---
title       : Working with dates
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=TRUE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Starting simple

```{r}
d1 = date()
d1
class(d1)
```

---

## Date class

```{r sysDate}
d2 = Sys.Date()
d2
class(d2)
```

---

## Formatting dates

`%d` = day as number (0-31), `%a` = abbreviated weekday,`%A` = unabbreviated weekday, `%m` = month (00-12), `%b` = abbreviated month,
`%B` = unabbrevidated month, `%y` = 2 digit year, `%Y` = four digit year

```{r ,dependson="sysDate"}
format(d2,"%a %b %d")
```

---

## Creating dates

```{r}
x = c("1jan1960", "2jan1960", "31mar1960", "30jul1960"); z = as.Date(x, "%d%b%Y")
z
z[1] - z[2]
as.numeric(z[1]-z[2])
```

---

## Converting to Julian 

```{r,dependson="sysDate"}
weekdays(d2)
months(d2)
julian(d2)
```

---

## Lubridate 

```{r lub}
library(lubridate); ymd("20140108")
mdy("08/04/2013")
dmy("03-04-2013")
```

[http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)

---

## Dealing with times

```{r ,dependson="lub"}
ymd_hms("2011-08-03 10:15:03")
ymd_hms("2011-08-03 10:15:03",tz="Pacific/Auckland")
?Sys.timezone
```

[http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)

---

## Some functions have slightly different syntax

```{r, dependson="lub"}
x = dmy(c("1jan2013", "2jan2013", "31mar2013", "30jul2013"))
wday(x[1])
wday(x[1],label=TRUE)
```

---

## Notes and further resources

* More information in this nice lubridate tutorial [http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/](http://www.r-statistics.com/2012/03/do-more-with-dates-and-times-in-r-with-lubridate-1-1-0/)
* The lubridate vignette is the same content [http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html](http://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)
* Ultimately you want your dates and times as class "Date" or the classes "POSIXct", "POSIXlt". For more information type `?POSIXlt`