---
title       : Representing data in R
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Important data types in R

__Classes__
* Character, Numeric, Integer, Logical

***********

__Objects__

* Vectors, Matrices, Data frames, Lists, Factors, Missing values

***********
__Operations__

* Subsetting, Logical subsetting

***********

_For more information_: 
* [Data Types](http://www.youtube.com/watch?v=5AQM-yUX9zg&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=5)


---

## Character

```{r}
firstName = "jeff"
class(firstName)
firstName
```

---

## Numeric
```{r}
heightCM = 188.2
class(heightCM)
heightCM
```

---

## Integer
```{r}
numberSons = 1L
class(numberSons)
numberSons
```

---

## Logical
```{r}
teachingCoursera = TRUE
class(teachingCoursera)
teachingCoursera
```

---
## Vectors
A set of values with the same class
```{r}
heights = c(188.2, 181.3, 193.4)
heights

firstNames = c("jeff","roger","andrew","brian")
firstNames

```

---

## Lists
A vector of values of possibly different classes
```{r}
vector1 = c(188.2, 181.3, 193.4)
vector2 = c("jeff","roger","andrew","brian")
myList = list(heights=vector1,firstNames=vector2)
myList

```

---

## Matrices
Vectors with multiple dimensions
```{r}
myMatrix = matrix(c(1,2,3,4),byrow=T,nrow=2)
myMatrix

```

---

## Data frames
Multiple vectors of possibly different classes, of the same length
```{r}
vector1 = c(188.2, 181.3, 193.4)
vector2 = c("jeff","roger","andrew","brian")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)
myDataFrame

```

---

## Data frames

```{r}
vector1 = c(188.2,181.3,193.4,192.3)
vector2 = c("jeff","roger","andrew","brian")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)
myDataFrame
```

---
## Factors
Qualitative variables that can be included in models

```{r}
smoker = c("yes","no","yes","yes")
smokerFactor = as.factor(smoker)
smokerFactor

```

---

## Missing values
In R they are usually coded NA

```{r}
vector1 = c(188.2,181.3,193.4,NA)
vector1
is.na(vector1)
```


---

## Subsetting

```{r}
vector1 = c(188.2,181.3,193.4,192.3)
vector2 = c("jeff","roger","andrew","brian")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)

vector1[1]
vector1[c(1,2,4)]

```

---

## Subsetting

```{r}
myDataFrame[1,1:2]
myDataFrame$firstNames

```
---

## Logical subsetting
```{r}
myDataFrame[myDataFrame$firstNames=="jeff",]
myDataFrame[myDataFrame$heights < 190,]
```

---

## Variable naming conventions

Variable names should be short, but descriptive. Here are some common styles

__Camel caps__
```{r}
myHeightCM = 188
```
__Underscore__
```{r}
my_height_cm = 188
```
__Dot separated__
```{r}
my.height.cm = 188
```

---

## Style guides

* [http://4dpiecharts.com/r-code-style-guide/](http://4dpiecharts.com/r-code-style-guide/)
* [http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html)
* [http://wiki.fhcrc.org/bioc/Coding_Standards](http://wiki.fhcrc.org/bioc/Coding_Standards)
