---
title       : EDA Case Study - Understanding Human Activity with Smart Phones
subtitle    : 
author      : Roger D. Peng, Associate Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


## Samsung Galaxy S3

<img class=center src=../../assets/img/samsung.png height=450>

[http://www.samsung.com/global/galaxys3/](http://www.samsung.com/global/galaxys3/)


---

## Samsung Data

<img class=center src=../../assets/img/ucisamsung.png height=450>

[http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)


---

## Slightly processed data

[Samsung data file]("https://dl.dropboxusercontent.com/u/7710864/courseraPublic/samsungData.rda")

```{r loaddata,tidy=TRUE}
load("data/samsungData.rda")
names(samsungData)[1:12]
table(samsungData$activity)
```

---

## Plotting average acceleration for first subject

```{r processData,fig.height=4,fig.width=8,tidy=TRUE}
par(mfrow=c(1, 2), mar = c(5, 4, 1, 1))
samsungData <- transform(samsungData, activity = factor(activity))
sub1 <- subset(samsungData, subject == 1)
plot(sub1[, 1], col = sub1$activity, ylab = names(sub1)[1])
plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[2])
legend("bottomright",legend=unique(sub1$activity),col=unique(sub1$activity), pch = 1)
```

---

## Clustering based just on average acceleration

<!-- ## source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R")  -->


```{r dependson="processData",fig.height=5,fig.width=8}
source("myplclust.R")
distanceMatrix <- dist(sub1[,1:3])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
```


---

## Plotting max acceleration for the first subject

```{r ,dependson="processData",fig.height=5,fig.width=10}
par(mfrow=c(1,2))
plot(sub1[,10],pch=19,col=sub1$activity,ylab=names(sub1)[10])
plot(sub1[,11],pch=19,col = sub1$activity,ylab=names(sub1)[11])
```

---

## Clustering based on maximum acceleration

```{r dependson="processData",fig.height=5,fig.width=10}
source("myplclust.R")
distanceMatrix <- dist(sub1[,10:12])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=unclass(sub1$activity))
```



---

## Singular Value Decomposition

```{r svdChunk,dependson="processData",fig.height=5,fig.width=10,cache=TRUE,tidy=TRUE}
svd1 = svd(scale(sub1[,-c(562,563)]))
par(mfrow=c(1,2))
plot(svd1$u[,1],col=sub1$activity,pch=19)
plot(svd1$u[,2],col=sub1$activity,pch=19)
```

---

## Find maximum contributor

```{r dependson="svdChunk",fig.height=5,fig.width=6,cache=TRUE,tidy=TRUE}
plot(svd1$v[,2],pch=19)
```


---

##  New clustering with maximum contributer

```{r dependson="svdChunk",fig.height=5,fig.width=8,cache=TRUE,tidy=TRUE}
maxContrib <- which.max(svd1$v[,2])
distanceMatrix <- dist(sub1[, c(10:12,maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=unclass(sub1$activity))                             
```


---

##  New clustering with maximum contributer

```{r dependson="svdChunk",fig.height=4.5,fig.width=4.5,cache=TRUE}
names(samsungData)[maxContrib]                          
```

---

##  K-means clustering (nstart=1, first try)

```{r kmeans1,dependson="processData",fig.height=4,fig.width=4}
kClust <- kmeans(sub1[,-c(562,563)],centers=6)
table(kClust$cluster,sub1$activity)
```



---

##  K-means clustering (nstart=1, second try)

```{r dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE,tidy=TRUE}
kClust <- kmeans(sub1[,-c(562,563)],centers=6,nstart=1)
table(kClust$cluster,sub1$activity)
```


---

##  K-means clustering (nstart=100, first try)

```{r dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE}
kClust <- kmeans(sub1[,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,sub1$activity)
```



---

##  K-means clustering (nstart=100, second try)

```{r kmeans100,dependson="kmeans1",fig.height=4,fig.width=4,cache=TRUE,tidy=TRUE}
kClust <- kmeans(sub1[,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,sub1$activity)
```

---

##  Cluster 1 Variable Centers (Laying)

```{r dependson="kmeans100",fig.height=4,fig.width=8,cache=FALSE,tidy=TRUE}
plot(kClust$center[1,1:10],pch=19,ylab="Cluster Center",xlab="")
```


---

##  Cluster 2 Variable Centers (Walking)

```{r dependson="kmeans100",fig.height=4,fig.width=8,cache=FALSE}
plot(kClust$center[4,1:10],pch=19,ylab="Cluster Center",xlab="")
```


