---
title       : Random forests
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', cache=TRUE, dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Random forests

1. Bootstrap samples
2. At each split, bootstrap variables
3. Grow multiple trees and vote

__Pros__:

1. Accuracy

__Cons__:

1. Speed
2. Interpretability
3. Overfitting


---

## Random forests

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/forests.png height=400>

[http://www.robots.ox.ac.uk/~az/lectures/ml/lect5.pdf](http://www.robots.ox.ac.uk/~az/lectures/ml/lect5.pdf)


---

## Iris data

```{r iris, cache=TRUE}
data(iris); library(ggplot2)
inTrain <- createDataPartition(y=iris$Species,
                              p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
```


---

## Random forests

```{r forestIris, dependson="irisData",fig.height=4,fig.width=4,cache=TRUE}
library(caret)
modFit <- train(Species~ .,data=training,method="rf",prox=TRUE)
modFit
```

---

## Getting a single tree

```{r , dependson="forestIris",fig.height=4,fig.width=4}
getTree(modFit$finalModel,k=2)
```

---

## Class "centers"

```{r centers, dependson="forestIris",fig.height=4,fig.width=4}
irisP <- classCenter(training[,c(3,4)], training$Species, modFit$finalModel$prox)
irisP <- as.data.frame(irisP); irisP$Species <- rownames(irisP)
p <- qplot(Petal.Width, Petal.Length, col=Species,data=training)
p + geom_point(aes(x=Petal.Width,y=Petal.Length,col=Species),size=5,shape=4,data=irisP)
```

---

## Predicting new values

```{r predForest, dependson="centers",fig.height=4,fig.width=4,cache=TRUE}
pred <- predict(modFit,testing); testing$predRight <- pred==testing$Species
table(pred,testing$Species)
```

---

## Predicting new values

```{r, dependson="predForest",fig.height=4,fig.width=4}
qplot(Petal.Width,Petal.Length,colour=predRight,data=testing,main="newdata Predictions")
```

---

## Notes and further resources

__Notes__:

* Random forests are usually one of the two top
performing algorithms along with boosting in prediction contests.
* Random forests are difficult to interpret but often very accurate. 
* Care should be taken to avoid overfitting (see [rfcv](http://cran.r-project.org/web/packages/randomForest/randomForest.pdf) funtion)


__Further resources__:

* [Random forests](http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm)
* [Random forest Wikipedia](http://en.wikipedia.org/wiki/Random_forest)
* [Elements of Statistical Learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
