---
title       : Introduction to regression
subtitle    : Regression
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, results='hide'}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
runif(1)
```

## A famous motivating example

<img class=center src=fig/galton.jpg height=150>

### (Perhaps surprisingly, this example is still relevant)

<img class=center src=fig/height.png height=150>

[http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html](http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html)

[Predicting height: the Victorian approach beats modern genomics](http://www.wired.com/wiredscience/2009/03/predicting-height-the-victorian-approach-beats-modern-genomics/)

---
## Recent simply statistics post
(Simply Statistics is a blog by Jeff Leek, Roger Peng and 
Rafael Irizarry, who wrote this post, link on the image)

<a href="http://simplystatistics.org/2013/01/28/data-supports-claim-that-if-kobe-stops-ball-hogging-the-lakers-will-win-more/">
<img class=center src=http://simplystatistics.org/wp-content/uploads/2013/01/kobelakers1-1024x1024.png height=250></img>
</a>

- "Data supports claim that if Kobe stops ball hogging the Lakers will win more"
- "Linear regression suggests that an increase of 1% in % of shots taken by Kobe results in a drop of 1.16 points (+/- 0.22)  in score differential."
- How was it done? Do you agree with the analysis? 






---
## Questions for this class
* Consider trying to answer the following kinds of questions:
  * To use the parents' heights to predict childrens' heights.
  * To try to find a parsimonious, easily described mean 
    relationship between parent and children's heights.
  * To investigate the variation in childrens' heights that appears 
  unrelated to parents' heights (residual variation).
  * To quantify what impact genotype information has beyond parental height in explaining child height.
  * To figure out how/whether and what assumptions are needed to
    generalize findings beyond the data in question.  
  * Why do children of very tall parents tend to be 
    tall, but a little shorter than their parents and why children of very short parents tend to be short, but a little taller than their parents? (This is a famous question called 'Regression to the mean'.)

---
## Galton's Data

* Let's look at the data first, used by Francis Galton in 1885. 
* Galton was a statistician who invented the term and concepts
  of regression and correlation, founded the journal Biometrika,
  and was the cousin of Charles Darwin.
* You may need to run `install.packages("UsingR")` if the `UsingR` library is not installed.
* Let's look at the marginal (parents disregarding children and children disregarding parents) distributions first. 
  * Parent distribution is all heterosexual couples.
  * Correction for gender via multiplying female heights by 1.08.
  * Overplotting is an issue from discretization.

---
```{r galton,fig.height=3.5,fig.width=8}
library(UsingR); data(galton); library(reshape); long <- melt(galton)
g <- ggplot(long, aes(x = value, fill = variable)) 
g <- g + geom_histogram(colour = "black", binwidth=1) 
g <- g + facet_grid(. ~ variable)
g
```

---
## Finding the middle via least squares
* Consider only the children's heights. 
  * How could one describe the "middle"?
  * One definition, let $Y_i$ be the height of child $i$ for $i = 1, \ldots, n = 928$, then define the middle as the value of $\mu$
  that minimizes $$\sum_{i=1}^n (Y_i - \mu)^2$$
* This is physical center of mass of the histrogram.
* You might have guessed that the answer $\mu = \bar Y$.


---
## Experiment
### Use R studio's manipulate to see what value of $\mu$ minimizes the sum of the squared deviations.

```
library(manipulate)
myHist <- function(mu){
    mse <- mean((galton$child - mu)^2)
    g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
    g <- g + geom_vline(xintercept = mu, size = 3)
    g <- g + ggtitle(paste("mu = ", mu, ", MSE = ", round(mse, 2), sep = ""))
    g
}
manipulate(myHist(mu), mu = slider(62, 74, step = 0.5))
```

---
## The least squares est. is the empirical mean
```{r , fig.height=4, fig.width=4, fig.align='center'}
g <- ggplot(galton, aes(x = child)) + geom_histogram(fill = "salmon", colour = "black", binwidth=1)
g <- g + geom_vline(xintercept = mean(galton$child), size = 3)
g
```

---
### The math (not required for the class) follows as:
$$ 
\begin{align} 
\sum_{i=1}^n (Y_i - \mu)^2 & = \
\sum_{i=1}^n (Y_i - \bar Y + \bar Y - \mu)^2 \\ 
& = \sum_{i=1}^n (Y_i - \bar Y)^2 + \
2 \sum_{i=1}^n (Y_i - \bar Y)  (\bar Y - \mu) +\
\sum_{i=1}^n (\bar Y - \mu)^2 \\
& = \sum_{i=1}^n (Y_i - \bar Y)^2 + \
2 (\bar Y - \mu) \sum_{i=1}^n (Y_i - \bar Y)  +\
\sum_{i=1}^n (\bar Y - \mu)^2 \\
& = \sum_{i=1}^n (Y_i - \bar Y)^2 + \
2 (\bar Y - \mu)  (\sum_{i=1}^n Y_i - n \bar Y) +\
\sum_{i=1}^n (\bar Y - \mu)^2 \\
& = \sum_{i=1}^n (Y_i - \bar Y)^2 + \sum_{i=1}^n (\bar Y - \mu)^2\\ 
& \geq \sum_{i=1}^n (Y_i - \bar Y)^2 \
\end{align} 
$$

---
## Comparing childrens' heights and their parents' heights

```{r, dependson="galton",fig.height=4,fig.width=4, fig.align='center'}
ggplot(galton, aes(x = parent, y = child)) + geom_point()
```

---
Size of point represents number of points at that (X, Y) combination (See the Rmd file for the code).

```{r freqGalton, dependson="galton",fig.height=6,fig.width=7,echo=FALSE}
library(dplyr)
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
g
```

---
## Regression through the origin
* Suppose that $X_i$ are the parents' heights.
* Consider picking the slope $\beta$ that minimizes $$\sum_{i=1}^n (Y_i - X_i \beta)^2$$
* This is exactly using the origin as a pivot point picking the
line that minimizes the sum of the squared vertical distances
of the points to the line
* Use R studio's  manipulate function to experiment
* Subtract the means so that the origin is the mean of the parent
and children's heights

---
```{r, echo = TRUE, eval = FALSE}
y <- galton$child - mean(galton$child)
x <- galton$parent - mean(galton$parent)
freqData <- as.data.frame(table(x, y))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
myPlot <- function(beta){
    g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
    g <- g  + scale_size(range = c(2, 20), guide = "none" )
    g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
    g <- g + geom_point(aes(colour=freq, size = freq))
    g <- g + scale_colour_gradient(low = "lightblue", high="white")                     
    g <- g + geom_abline(intercept = 0, slope = beta, size = 3)
    mse <- mean( (y - beta * x) ^2 )
    g <- g + ggtitle(paste("beta = ", beta, "mse = ", round(mse, 3)))
    g
}
manipulate(myPlot(beta), beta = slider(0.6, 1.2, step = 0.02))
```

---
## The solution 
### In the next few lectures we'll talk about why this is the solution
```{r}
lm(I(child - mean(child))~ I(parent - mean(parent)) - 1, data = galton)
```

---
```{r, fig.height=6,fig.width=7,echo=FALSE}
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
lm1 <- lm(galton$child ~ galton$parent)
g <- g + geom_abline(intercept = coef(lm1)[1], slope = coef(lm1)[2], size = 3, colour = grey(.5))
g
```

