---
title       : Count outcomes, Poisson GLMs
subtitle    : Regression Models
author      : Brian Caffo, Jeffrey Leek, Roger Peng 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Many data take the form of counts
  * Calls to a call center
  * Number of flu cases in an area
  * Number of cars that cross a bridge
* Data may also be in the form of rates
  * Percent of children passing a test
  * Percent of hits to a website from a country
* Linear regression with transformation is an option

---

## Poisson distribution
- The Poisson distribution is a useful model for counts and rates
- Here a rate is count per some monitoring time
- Some examples uses of the Poisson distribution
    - Modeling web traffic hits
    - Incidence rates
    - Approximating binomial probabilities with small $p$ and large $n$
    - Analyzing contigency table data

---
## The Poisson mass function
- $X \sim Poisson(t\lambda)$ if
$$
P(X = x) = \frac{(t\lambda)^x e^{-t\lambda}}{x!}
$$
For $x = 0, 1, \ldots$.
- The mean of the Poisson is $E[X] = t\lambda$, thus $E[X / t] = \lambda$
- The variance of the Poisson is $Var(X) = t\lambda$.
- The Poisson tends to a normal as $t\lambda$ gets large.

---

```{r simPois,fig.height=4,fig.width=8, cache=TRUE}
par(mfrow = c(1, 3))
plot(0 : 10, dpois(0 : 10, lambda = 2), type = "h", frame = FALSE)
plot(0 : 20, dpois(0 : 20, lambda = 10), type = "h", frame = FALSE)
plot(0 : 200, dpois(0 : 200, lambda = 100), type = "h", frame = FALSE) 
```

---

## Poisson distribution
### Sort of, showing that the mean and variance are equal
```{r}
x <- 0 : 10000; lambda = 3
mu <- sum(x * dpois(x, lambda = lambda))
sigmasq <- sum((x - mu)^2 * dpois(x, lambda = lambda))
c(mu, sigmasq)
```

---

## Example: Leek Group Website Traffic
* Consider the daily counts to Jeff Leek's web site

[http://biostat.jhsph.edu/~jleek/](http://biostat.jhsph.edu/~jleek/)

* Since the unit of time is always one day, set $t = 1$ and then
the Poisson mean is interpretted as web hits per day. (If we set $t = 24$, it would
be web hits per hour).

---

## Website data

```{r leekLoad,cache=TRUE}
download.file("https://dl.dropboxusercontent.com/u/7710864/data/gaData.rda",destfile="./data/gaData.rda",method="curl")
load("./data/gaData.rda")
gaData$julian <- julian(gaData$date)
head(gaData)
```

[http://skardhamar.github.com/rga/](http://skardhamar.github.com/rga/)


---

## Plot data

```{r, dependson="leekLoad",fig.height=4.5,fig.width=4.5}
plot(gaData$julian,gaData$visits,pch=19,col="darkgrey",xlab="Julian",ylab="Visits")
```


---

## Linear regression

$$ NH_i = b_0 + b_1 JD_i + e_i $$

$NH_i$ - number of hits to the website

$JD_i$ - day of the year (Julian day)

$b_0$ - number of hits on Julian day 0 (1970-01-01)

$b_1$ - increase in number of hits per unit day

$e_i$ - variation due to everything we didn't measure


---

## Linear regression line

```{r linReg, dependson="leekLoad",fig.height=4,fig.width=4, cache=TRUE}
plot(gaData$julian,gaData$visits,pch=19,col="darkgrey",xlab="Julian",ylab="Visits")
lm1 <- lm(gaData$visits ~ gaData$julian)
abline(lm1,col="red",lwd=3)
```
---

## Aside, taking the log of the outcome
- Taking the natural log of the outcome has a specific interpretation.
- Consider the model

$$ \log(NH_i) = b_0 + b_1 JD_i + e_i $$

$NH_i$ - number of hits to the website

$JD_i$ - day of the year (Julian day)

$b_0$ - log number of hits on Julian day 0 (1970-01-01)

$b_1$ - increase in log number of hits per unit day

$e_i$ - variation due to everything we didn't measure

---
## Exponentiating coefficients
- $e^{E[\log(Y)]}$ geometric mean of $Y$. 
    - With no covariates, this is estimated by $e^{\frac{1}{n}\sum_{i=1}^n \log(y_i)} = (\prod_{i=1}^n y_i)^{1/n}$
- When you take the natural log of outcomes and fit a regression model, your exponentiated coefficients
estimate things about geometric means.
- $e^{\beta_0}$ estimated geometric mean hits on day 0
- $e^{\beta_1}$ estimated relative increase or decrease in geometric mean hits per day
- There's a problem with logs with you have zero counts, adding a constant works
```{r}
round(exp(coef(lm(I(log(gaData$visits + 1)) ~ gaData$julian))), 5)
```

---

## Linear vs. Poisson regression

__Linear__

$$ NH_i = b_0 + b_1 JD_i + e_i $$

or

$$ E[NH_i | JD_i, b_0, b_1] = b_0 + b_1 JD_i$$

__Poisson/log-linear__

$$ \log\left(E[NH_i | JD_i, b_0, b_1]\right) = b_0 + b_1 JD_i $$

or

$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 + b_1 JD_i\right) $$


---

## Multiplicative differences

<br><br>
$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 + b_1 JD_i\right) $$

<br><br>

$$ E[NH_i | JD_i, b_0, b_1] = \exp\left(b_0 \right)\exp\left(b_1 JD_i\right) $$

<br><br>

If $JD_i$ is increased by one unit, $E[NH_i | JD_i, b_0, b_1]$ is multiplied by $\exp\left(b_1\right)$

---

## Poisson regression in R

```{r poisReg, dependson="linReg",fig.height=4.5,fig.width=4.5, cache=TRUE}
plot(gaData$julian,gaData$visits,pch=19,col="darkgrey",xlab="Julian",ylab="Visits")
glm1 <- glm(gaData$visits ~ gaData$julian,family="poisson")
abline(lm1,col="red",lwd=3); lines(gaData$julian,glm1$fitted,col="blue",lwd=3)
```


---

## Mean-variance relationship?

```{r, dependson="poisReg",fig.height=4.5,fig.width=4.5}
plot(glm1$fitted,glm1$residuals,pch=19,col="grey",ylab="Residuals",xlab="Fitted")
```

---

## Model agnostic standard errors 

```{r agnostic}
library(sandwich)
confint.agnostic <- function (object, parm, level = 0.95, ...)
{
    cf <- coef(object); pnames <- names(cf)
    if (missing(parm))
        parm <- pnames
    else if (is.numeric(parm))
        parm <- pnames[parm]
    a <- (1 - level)/2; a <- c(a, 1 - a)
    pct <- stats:::format.perc(a, 3)
    fac <- qnorm(a)
    ci <- array(NA, dim = c(length(parm), 2L), dimnames = list(parm,
                                                               pct))
    ses <- sqrt(diag(sandwich::vcovHC(object)))[parm]
    ci[] <- cf[parm] + ses %o% fac
    ci
}
```
[http://stackoverflow.com/questions/3817182/vcovhc-and-confidence-interval](http://stackoverflow.com/questions/3817182/vcovhc-and-confidence-interval)

---

## Estimating confidence intervals

```{r}
confint(glm1)
confint.agnostic(glm1)
```


---

## Rates 


<br><br>


$$ E[NHSS_i | JD_i, b_0, b_1]/NH_i = \exp\left(b_0 + b_1 JD_i\right) $$

<br><br>

$$ \log\left(E[NHSS_i | JD_i, b_0, b_1]\right) - \log(NH_i)  =  b_0 + b_1 JD_i $$

<br><br>

$$ \log\left(E[NHSS_i | JD_i, b_0, b_1]\right) = \log(NH_i) + b_0 + b_1 JD_i $$

---

## Fitting rates in R 

```{r ratesFit,dependson="agnostic", cache=TRUE,fig.height=4,fig.width=4}
glm2 <- glm(gaData$simplystats ~ julian(gaData$date),offset=log(visits+1),
            family="poisson",data=gaData)
plot(julian(gaData$date),glm2$fitted,col="blue",pch=19,xlab="Date",ylab="Fitted Counts")
points(julian(gaData$date),glm1$fitted,col="red",pch=19)
```

---

## Fitting rates in R 

```{r,dependson="ratesFit",fig.height=4,fig.width=4}
glm2 <- glm(gaData$simplystats ~ julian(gaData$date),offset=log(visits+1),
            family="poisson",data=gaData)
plot(julian(gaData$date),gaData$simplystats/(gaData$visits+1),col="grey",xlab="Date",
     ylab="Fitted Rates",pch=19)
lines(julian(gaData$date),glm2$fitted/(gaData$visits+1),col="blue",lwd=3)
```

---

## More information

* [Log-linear models and multiway tables](http://ww2.coastal.edu/kingw/statistics/R-tutorials/loglin.html)
* [Wikipedia on Poisson regression](http://en.wikipedia.org/wiki/Poisson_regression), [Wikipedia on overdispersion](http://en.wikipedia.org/wiki/Overdispersion)
* [Regression models for count data in R](http://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf)
* [pscl package](http://cran.r-project.org/web/packages/pscl/index.html) - the function _zeroinfl_ fits zero inflated models. 
