---
title       : Predicting with regression, multiple covariates
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Example: predicting wages

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/wages.jpg height=350>

Image Credit [http://www.cahs-media.org/the-high-cost-of-low-wages](http://www.cahs-media.org/the-high-cost-of-low-wages)

Data from: [ISLR package](http://cran.r-project.org/web/packages/ISLR) from the book: [Introduction to statistical learning](http://www-bcf.usc.edu/~gareth/ISL/)



---

## Example: Wage data

```{r loadData,cache=TRUE}
library(ISLR); library(ggplot2); library(caret);
data(Wage); Wage <- subset(Wage,select=-c(logwage))
summary(Wage)
```



---

## Get training/test sets

```{r trainingTest,dependson="loadData",cache=TRUE}
inTrain <- createDataPartition(y=Wage$wage,
                              p=0.7, list=FALSE)
training <- Wage[inTrain,]; testing <- Wage[-inTrain,]
dim(training); dim(testing)
```



---

## Feature plot

```{r ,dependson="trainingTest",fig.height=4,fig.width=4}
featurePlot(x=training[,c("age","education","jobclass")],
            y = training$wage,
            plot="pairs")
```


---

## Plot age versus wage


```{r ,dependson="trainingTest",fig.height=4,fig.width=6}
qplot(age,wage,data=training)
```


---

## Plot age versus wage colour by jobclass


```{r ,dependson="trainingTest",fig.height=4,fig.width=6}
qplot(age,wage,colour=jobclass,data=training)
```


---

## Plot age versus wage colour by education


```{r ,dependson="trainingTest",fig.height=4,fig.width=6}
qplot(age,wage,colour=education,data=training)
```

---

## Fit a linear model 

$$ ED_i = b_0 + b_1 age + b_2 I(Jobclass_i="Information") + \sum_{k=1}^4 \gamma_k I(education_i= level k) $$

```{r modelFit,dependson="trainingTest", cache=TRUE,fig.height=4,fig.width=4}
modFit<- train(wage ~ age + jobclass + education,
               method = "lm",data=training)
finMod <- modFit$finalModel
print(modFit)
```

Education levels: 1 = HS Grad, 2 = Some College, 3 = College Grad, 4 = Advanced Degree

---

## Diagnostics

```{r,dependson="modelFit",fig.height=5,fig.width=5}
plot(finMod,1,pch=19,cex=0.5,col="#00000010")
```


---

## Color by variables not used in the model 

```{r,dependson="modelFit",fig.height=4,fig.width=6}
qplot(finMod$fitted,finMod$residuals,colour=race,data=training)
```

---

## Plot by index

```{r,dependson="modelFit",fig.height=5,fig.width=5}
plot(finMod$residuals,pch=19)
```


---

## Predicted versus truth in test set

```{r predictions, dependson="modelFit",fig.height=4,fig.width=6}
pred <- predict(modFit, testing)
qplot(wage,pred,colour=year,data=testing)
```

---

## If you want to use all covariates

```{r allCov,dependson="trainingTest",fig.height=4,fig.width=4,warning=FALSE}
modFitAll<- train(wage ~ .,data=training,method="lm")
pred <- predict(modFitAll, testing)
qplot(wage,pred,data=testing)
```


---

## Notes and further reading

* Often useful in combination with other models 
* [Elements of statistical learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
* [Modern applied statistics with S](http://www.amazon.com/Modern-Applied-Statistics-W-N-Venables/dp/0387954570)
* [Introduction to statistical learning](http://www-bcf.usc.edu/~gareth/ISL/)