---
title       : Inference in regression
subtitle    : 
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, results='hide'}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
runif(1)
```
## Recall our model and fitted values
* Consider the model
$$
Y_i = \beta_0 + \beta_1 X_i + \epsilon_i
$$
* $\epsilon \sim N(0, \sigma^2)$. 
* We assume that the true model is known.
* We assume that you've seen confidence intervals and hypothesis tests before.
* $\hat \beta_0 = \bar Y - \hat \beta_1 \bar X$
* $\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)}$.

---
## Review
* Statistics like $\frac{\hat \theta - \theta}{\hat \sigma_{\hat \theta}}$ often have the following properties.
    1. Is normally distributed and has a finite sample Student's T distribution if the  variance is replaced with a sample estimate (under normality assumptions).
    3. Can be used to test $H_0 : \theta = \theta_0$ versus $H_a : \theta >, <, \neq \theta_0$.
    4. Can be used to create a confidence interval for $\theta$ via $\hat \theta \pm Q_{1-\alpha/2} \hat \sigma_{\hat \theta}$
    where $Q_{1-\alpha/2}$ is the relevant quantile from either a normal or T distribution.
* In the case of regression with iid sampling assumptions and normal errors, our inferences will follow
very similarily to what you saw in your inference class.
* We won't cover asymptotics for regression analysis, but suffice it to say that under assumptions 
on the ways in which the $X$ values are collected, the iid sampling model, and mean model, 
the normal results hold to create intervals and confidence intervals

---
## Results
* $\sigma_{\hat \beta_1}^2 = Var(\hat \beta_1) = \sigma^2 / \sum_{i=1}^n (X_i - \bar X)^2$
* $\sigma_{\hat \beta_0}^2 = Var(\hat \beta_0)  = \left(\frac{1}{n} + \frac{\bar X^2}{\sum_{i=1}^n (X_i - \bar X)^2 }\right)\sigma^2$
* In practice, $\sigma$ is replaced by its estimate.
* It's probably not surprising that under iid Gaussian errors
$$
\frac{\hat \beta_j - \beta_j}{\hat \sigma_{\hat \beta_j}}
$$
follows a $t$ distribution with $n-2$ degrees of freedom and a normal distribution for large $n$.
* This can be used to create confidence intervals and perform
hypothesis tests.

---
## Example diamond data set
```{r}
library(UsingR); data(diamond)
y <- diamond$price; x <- diamond$carat; n <- length(y)
beta1 <- cor(y, x) * sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
e <- y - beta0 - beta1 * x
sigma <- sqrt(sum(e^2) / (n-2)) 
ssx <- sum((x - mean(x))^2)
seBeta0 <- (1 / n + mean(x) ^ 2 / ssx) ^ .5 * sigma 
seBeta1 <- sigma / sqrt(ssx)
tBeta0 <- beta0 / seBeta0; tBeta1 <- beta1 / seBeta1
pBeta0 <- 2 * pt(abs(tBeta0), df = n - 2, lower.tail = FALSE)
pBeta1 <- 2 * pt(abs(tBeta1), df = n - 2, lower.tail = FALSE)
coefTable <- rbind(c(beta0, seBeta0, tBeta0, pBeta0), c(beta1, seBeta1, tBeta1, pBeta1))
colnames(coefTable) <- c("Estimate", "Std. Error", "t value", "P(>|t|)")
rownames(coefTable) <- c("(Intercept)", "x")
```

---
## Example continued

```{r}
coefTable
fit <- lm(y ~ x); 
summary(fit)$coefficients
```

---
## Getting a confidence interval
```{r}
sumCoef <- summary(fit)$coefficients
sumCoef[1,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[1, 2]
(sumCoef[2,1] + c(-1, 1) * qt(.975, df = fit$df) * sumCoef[2, 2]) / 10
```
With 95% confidence, we estimate that a 0.1 carat increase in
diamond size results in a `r round((sumCoef[2,1] - qt(.975, df = fit$df) * sumCoef[2, 2]) / 10, 1)` to `r round((sumCoef[2,1] + qt(.975, df = fit$df) * sumCoef[2, 2]) / 10, 1)` increase in price in (Singapore) dollars.

---
## Prediction of outcomes
* Consider predicting $Y$ at a value of $X$
  * Predicting the price of a diamond given the carat
  * Predicting the height of a child given the height of the parents
* The obvious estimate for prediction at point $x_0$ is 
$$
\hat \beta_0 + \hat \beta_1 x_0
$$
* A standard error is needed to create a prediction interval.
* There's a distinction between intervals for the regression
  line at point $x_0$ and the prediction of what a $y$ would be
  at point $x_0$. 
* Line at $x_0$ se, $\hat \sigma\sqrt{\frac{1}{n} +  \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}$
* Prediction interval se at $x_0$, $\hat \sigma\sqrt{1 + \frac{1}{n} + \frac{(x_0 - \bar X)^2}{\sum_{i=1}^n (X_i - \bar X)^2}}$


---
## Plotting the prediction intervals
```{r, fig.height=5, fig.width==5, echo = FALSE, results='hide'}
library(ggplot2)
newx = data.frame(x = seq(min(x), max(x), length = 100))
p1 = data.frame(predict(fit, newdata= newx,interval = ("confidence")))
p2 = data.frame(predict(fit, newdata = newx,interval = ("prediction")))
p1$interval = "confidence"
p2$interval = "prediction"
p1$x = newx$x
p2$x = newx$x
dat = rbind(p1, p2)
names(dat)[1] = "y"

g = ggplot(dat, aes(x = x, y = y))
g = g + geom_ribbon(aes(ymin = lwr, ymax = upr, fill = interval), alpha = 0.2) 
g = g + geom_line()
g = g + geom_point(data = data.frame(x = x, y=y), aes(x = x, y = y), size = 4)
g
```

---
## Discussion
* Both intervals have varying widths.
  * Least width at the mean of the Xs.
* We are quite confident in the regression line, so that 
  interval is very narrow.
  * If we knew $\beta_0$ and $\beta_1$ this interval would have zero width.
* The prediction interval must incorporate the variabilibity
  in the data around the line.
  * Even if we knew $\beta_0$ and $\beta_1$ this interval would still have width.

---

## In R
```{r, fig.height=5, fig.width=5, echo=FALSE,results='hide'}
newdata <- data.frame(x = xVals)
p1 <- predict(fit, newdata, interval = ("confidence"))
p2 <- predict(fit, newdata, interval = ("prediction"))
plot(x, y, frame=FALSE,xlab="Carat",ylab="Dollars",pch=21,col="black", bg="lightblue", cex=2)
abline(fit, lwd = 2)
lines(xVals, p1[,2]); lines(xVals, p1[,3])
lines(xVals, p2[,2]); lines(xVals, p2[,3])
```
  
  
  
  
  




