% Managing Data Frames with `dplyr`
%

```{r, echo=FALSE, results="hide"}
options(width = 50)
```

# dplyr

The data frame is a key data structure in statistics and in R.

* There is one observation per row

* Each column represents a variable or measure or characteristic

* Primary implementation that you will use is the default R
  implementation

* Other implementations, particularly relational databases systems


# dplyr

* Developed by Hadley Wickham of RStudio

* An optimized and distilled version of `plyr` package (also by Hadley)

* Does not provide any "new" functionality per se, but **greatly**
  simplifies existing functionality in R

* Provides a "grammar" (in particular, verbs) for data manipulation

* Is **very** fast, as many key operations are coded in C++


# dplyr Verbs

* `select`: return a subset of the columns of a data frame

* `filter`: extract a subset of rows from a data frame based on
  logical conditions

* `arrange`: reorder rows of a data frame


* `rename`: rename variables in a data frame

* `mutate`: add new variables/columns or transform existing variables

* `summarise` / `summarize`: generate summary statistics of different
  variables in the data frame, possibly within strata

There is also a handy `print` method that prevents you from printing a
lot of data to the console.



# dplyr Properties

* The first argument is a data frame.

* The subsequent arguments describe what to do with it, and you can
  refer to columns in the data frame directly without using the $
  operator (just use the names).

* The result is a new data frame

* Data frames must be properly formatted and annotated for this to all
  be useful


# Load the `dplyr` package


This step is important!

```{r}
library(dplyr)
```


# `select`

```{r}
chicago <- readRDS("chicago.rds")
dim(chicago)
head(select(chicago, 1:5))
```


# `select`

```{r}
names(chicago)[1:3]
head(select(chicago, city:dptp))
```

# `select`

In dplyr you can do

```{r,eval=FALSE}
head(select(chicago, -(city:dptp)))
```

Equivalent base R

```{r,eval=FALSE}
i <- match("city", names(chicago))
j <- match("dptp", names(chicago))
head(chicago[, -(i:j)])
```



# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30)
head(select(chic.f, 1:3, pm25tmean2), 10)
```

# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(select(chic.f, 1:3, pm25tmean2, tmpd), 10)
```


# `arrange`

Reordering rows of a data frame (while preserving corresponding order
of other columns) is normally a pain to do in R.

```{r}
chicago <- arrange(chicago, date)
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```

# `arrange`

Columns can be arranged in descending order too.

```{r}
chicago <- arrange(chicago, desc(date))
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```


# `rename`

Renaming a variable in a data frame in R is surprising hard to do!

```{r,tidy=FALSE}
head(chicago[, 1:5], 3)
chicago <- rename(chicago, dewpoint = dptp, 
                  pm25 = pm25tmean2)
head(chicago[, 1:5], 3)
```


# `mutate`

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  pm25detrend=pm25-mean(pm25, na.rm=TRUE))
head(select(chicago, pm25, pm25detrend))
```

# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  tempcat = factor(1 * (tmpd > 80), 
                                   labels = c("cold", "hot")))
hotcold <- group_by(chicago, tempcat)
summarize(hotcold, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2), 
          no2 = median(no2tmean2))
```


# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  year = as.POSIXlt(date)$year + 1900)
years <- group_by(chicago, year)
summarize(years, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago$year <- NULL  ## Can't use mutate to create an existing variable
```


# `%>%`

```{r,tidy=FALSE,eval=FALSE}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) 
	%>% group_by(month) 
	%>% summarize(pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago %>% mutate(month = as.POSIXlt(date)$mon + 1) %>% group_by(month) %>% 
summarize(pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2, na.rm = TRUE), no2 = median(no2tmean2, na.rm = TRUE))

```


# dplyr

Once you learn the dplyr "grammar" there are a few additional benefits

* dplyr can work with other data frame "backends"

* `data.table` for large fast tables

* SQL interface for relational databases via the DBI package


