---
title       : Reading JSON
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=TRUE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```



## JSON

* Javascript Object Notation
* Lightweight data storage
* Common format for data from application programming interfaces (APIs)
* Similar structure to XML but different syntax/format
* Data stored as
  * Numbers (double)
  * Strings (double quoted)
  * Boolean (_true_ or _false_)
  * Array (ordered, comma separated enclosed in square brackets _[]_)
  * Object (unorderd, comma separated collection of key:value pairs in curley brackets _{}_)


[http://en.wikipedia.org/wiki/JSON](http://en.wikipedia.org/wiki/JSON)


---

## Example JSON file

<img class=center src=../../assets/img/03_ObtainingData/githubjson.png height=450>


---

## Reading data from JSON {jsonlite package}

```{r readJSON}
library(jsonlite)
jsonData <- fromJSON("https://api.github.com/users/jtleek/repos")
names(jsonData)
jsonData$name
```


---

## Nested objects in JSON

```{r,dependson="readJSON"}
names(jsonData$owner)
jsonData$owner$login
```


---

## Writing data frames to JSON

```{r writeJSON}
myjson <- toJSON(iris, pretty=TRUE)
cat(myjson)
```

[http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/](http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/)


---

## Convert back to JSON

```{r,dependson="writeJSON"}
iris2 <- fromJSON(myjson)
head(iris2)
```


[http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/](http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/)

---

## Further resources

* [http://www.json.org/](http://www.json.org/)
* A good tutorial on jsonlite - [http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/](http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/)
* [jsonlite vignette](http://cran.r-project.org/web/packages/jsonlite/vignettes/json-mapping.pdf)