---
title       : Exploratory graphs
subtitle    : 
author      : Roger D. Peng, Associate Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache = T, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Why do we use graphs in data analysis? 

* To understand data properties
* To find patterns in data
* To suggest modeling strategies
* To "debug" analyses
* To communicate results

---

## Exploratory graphs

* <redtext>To understand data properties</redtext>
* <redtext>To find patterns in data</redtext>
* <redtext>To suggest modeling strategies</redtext>
* <redtext>To "debug" analyses</redtext>
* To communicate results

---

## Characteristics of exploratory graphs

* They are made quickly
* A large number are made
* The goal is for personal understanding
* Axes/legends are generally cleaned up
* Color/size are primarily used for information

---

## Background - perceptual tasks

<img class=center src=../../assets/img/perceptual.png height=450>

[Graphical perception: Theory, Experimentation, and Applications to the Development of Graphical Models](http://www.jstor.org/discover/10.2307/2288400?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101742782357)

---

## Position versus length

<img class=center src=../../assets/img/barcharts.png height=250>

[Graphical perception: Theory, Experimentation, and Applications to the Development of Graphical Models](http://www.jstor.org/discover/10.2307/2288400?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101742782357)


---

## Position versus length - results

<img class=center src=../../assets/img/poslength.png height=250>


[Graphical perception: Theory, Experimentation, and Applications to the Development of Graphical Models](http://www.jstor.org/discover/10.2307/2288400?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101742782357)

---

## Position versus angle

<img class=center src=../../assets/img/piechart.png height=300>


[Graphical perception: Theory, Experimentation, and Applications to the Development of Graphical Models](http://www.jstor.org/discover/10.2307/2288400?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101742782357)


---

## Position versus angle - results


<img class=center src=../../assets/img/posangle.png height=200>

<br><br>
[Graphical perception: Theory, Experimentation, and Applications to the Development of Graphical Models](http://www.jstor.org/discover/10.2307/2288400?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101742782357)


---

## More experimental results

<img class=center src=../../assets/img/bigexp.png height=450>

[Graphical Perception and Graphical Methods for Analyzing Scientific Data](http://www.sciencemag.org/content/229/4716/828.refs)

---

## Summary

* Use common scales when possible
* When possible use position comparisons
* Angle comparisons are frequently hard to interpret (no piecharts!)
* No 3-D barcharts. Please.


---

## Housing data


<img class=center src=../../assets/img/acs.png height=350>

<br>

```{r downloadData, echo=FALSE, cache=TRUE}
if(!file.exists("data/ss06pid.csv")) {
        download.file("https://dl.dropboxusercontent.com/u/7710864/data/csv_hid/ss06pid.csv",destfile="./data/ss06pid.csv",method="curl")
}
```


```{r loadData,dependson="downloadData"}
pData <- read.csv("./data/ss06pid.csv")
```


--- 

## Boxplots

Important parameters: _col_,_varwidth_,_names_,_horizontal_

```{r , dependson="loadData",fig.height=4,fig.width=4}
boxplot(pData$AGEP, col="blue")
```

--- 

## Boxplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
pData <- transform(pData, DDRS = factor(DDRS))
boxplot(AGEP ~ DDRS, data = pData, col = "blue")
```


--- 

## Boxplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
boxplot(AGEP ~ DDRS, data = pData, col = c("blue", "orange"), 
        names = c("yes", "no"), varwidth = TRUE)
```



--- 

## Barplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
barplot(table(pData$CIT), col = "blue")
```



--- 

## Histograms

Important parameters: _breaks_,_freq_,_col_,_xlab_,_ylab_, _xlim, _ylim_ ,_main_

```{r , dependson="loadData",fig.height=4,fig.width=4}
hist(pData$AGEP, col = "blue")
```

--- 

## Histograms

```{r , dependson="loadData",fig.height=5,fig.width=5}
hist(pData$AGEP, col = "blue", breaks = 100, main = "Age")
```


--- 

## Density plots

Important parameters (to plot): _col_,_lwd_,_xlab_,_ylab_,_xlim_,_ylim_

```{r , dependson="loadData",fig.height=4,fig.width=4}
dens <- density(pData$AGEP)
plot(dens, lwd = 3, col = "blue")
```


--- 

## Density plots - multiple distributions

```{r , dependson="loadData",fig.height=4,fig.width=4}
dens <- density(pData$AGEP)
densMales <- density(pData$AGEP[which(pData$SEX == 1)])
plot(dens, lwd = 3, col = "blue")
lines(densMales, lwd = 3, col = "orange")
```



