---
title       : Power
subtitle    : Statistical Inference
author      : Brian Caffo, Jeff Leek, Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Power
- Power is the probability of rejecting the null hypothesis when it is false
- Ergo, power (as its name would suggest) is a good thing; you want more power
- A type II error (a bad thing, as its name would suggest) is failing to reject the null hypothesis when it's false; the probability of a type II error is usually called $\beta$
- Note Power  $= 1 - \beta$

---
## Notes
- Consider our previous example involving RDI
- $H_0: \mu = 30$ versus $H_a: \mu > 30$
- Then power is 
$$P\left(\frac{\bar X - 30}{s /\sqrt{n}} > t_{1-\alpha,n-1} ~|~ \mu = \mu_a \right)$$
- Note that this is a function that depends on the specific value of $\mu_a$!
- Notice as $\mu_a$ approaches $30$ the power approaches $\alpha$


---
## Calculating power for Gaussian data
Assume that $n$ is large and that we know $\sigma$
$$
\begin{align}
1 -\beta & = 
P\left(\frac{\bar X - 30}{\sigma /\sqrt{n}} > z_{1-\alpha} ~|~ \mu = \mu_a \right)\\
& = P\left(\frac{\bar X - \mu_a + \mu_a - 30}{\sigma /\sqrt{n}} > z_{1-\alpha} ~|~ \mu = \mu_a \right)\\ \\
& = P\left(\frac{\bar X - \mu_a}{\sigma /\sqrt{n}} > z_{1-\alpha} - \frac{\mu_a - 30}{\sigma /\sqrt{n}} ~|~ \mu = \mu_a \right)\\ \\
& = P\left(Z > z_{1-\alpha} - \frac{\mu_a - 30}{\sigma /\sqrt{n}} ~|~ \mu = \mu_a \right)\\ \\
\end{align}
$$

---
## Example continued
-  Suppose that we wanted to detect a increase in mean RDI
  of at least 2 events / hour (above 30). 
- Assume normality and that the sample in question will have a standard deviation of $4$;
- What would be the power if we took a sample size of $16$?
  - $Z_{1-\alpha} = 1.645$ 
  - $\frac{\mu_a - 30}{\sigma /\sqrt{n}} = 2 / (4 /\sqrt{16}) = 2$ 
  - $P(Z > 1.645 - 2) = P(Z > -0.355) = 64\%$
```{r}
pnorm(-0.355, lower.tail = FALSE)
```

---
## Note 
- Consider $H_0 : \mu = \mu_0$ and $H_a : \mu > \mu_0$ with $\mu = \mu_a$ under $H_a$.
- Under $H_0$ the statistic $Z = \frac{\sqrt{n}(\bar X - \mu_0)}{\sigma}$ is $N(0, 1)$
- Under $H_a$ $Z$ is $N\left( \frac{\sqrt{n}(\mu_a - \mu_0)}{\sigma}, 1\right)$
- We reject if $Z > Z_{1-\alpha}$

```
sigma <- 10; mu_0 = 0; mu_a = 2; n <- 100; alpha = .05
plot(c(-3, 6),c(0, dnorm(0)), type = "n", frame = FALSE, xlab = "Z value", ylab = "")
xvals <- seq(-3, 6, length = 1000)
lines(xvals, dnorm(xvals), type = "l", lwd = 3)
lines(xvals, dnorm(xvals, mean = sqrt(n) * (mu_a - mu_0) / sigma), lwd =3)
abline(v = qnorm(1 - alpha))
```

---
```{r, fig.height=5, fig.width=5, echo = FALSE}
sigma <- 10; mu_0 = 0; mu_a = 2; n <- 100; alpha = .05
plot(c(-3, 6),c(0, dnorm(0)), type = "n", frame = FALSE, xlab = "Z value", ylab = "")
xvals <- seq(-3, 6, length = 1000)
lines(xvals, dnorm(xvals), type = "l", lwd = 3)
lines(xvals, dnorm(xvals, mean = sqrt(n) * (mu_a - mu_0) / sigma), lwd =3)
abline(v = qnorm(1 - alpha))
```


---
## Question
- When testing $H_a : \mu > \mu_0$, notice if power is $1 - \beta$, then 
$$1 - \beta = P\left(Z > z_{1-\alpha} - \frac{\mu_a - \mu_0}{\sigma /\sqrt{n}} ~|~ \mu = \mu_a \right) = P(Z > z_{\beta})$$
- This yields the equation
$$z_{1-\alpha} - \frac{\sqrt{n}(\mu_a - \mu_0)}{\sigma} = z_{\beta}$$
- Unknowns: $\mu_a$, $\sigma$, $n$, $\beta$
- Knowns: $\mu_0$, $\alpha$
- Specify any 3 of the unknowns and you can solve for the remainder

---
## Notes
- The calculation for $H_a:\mu < \mu_0$ is similar
- For $H_a: \mu \neq \mu_0$ calculate the one sided power using
  $\alpha / 2$ (this is only approximately right, it excludes the probability of
  getting a large TS in the opposite direction of the truth)
- Power goes up as $\alpha$ gets larger
- Power of a one sided test is greater than the power of the
  associated two sided test
- Power goes up as $\mu_1$ gets further away from $\mu_0$
- Power goes up as $n$ goes up
- Power doesn't need $\mu_a$, $\sigma$ and $n$, instead only $\frac{\sqrt{n}(\mu_a - \mu_0)}{\sigma}$
  - The quantity $\frac{\mu_a - \mu_0}{\sigma}$ is called the effect size, the difference in the means in standard deviation units.
  - Being unit free, it has some hope of interpretability across settings

---
## T-test power
-  Consider calculating power for a Gossett's $T$ test for our example
-  The power is
  $$
  P\left(\frac{\bar X - \mu_0}{S /\sqrt{n}} > t_{1-\alpha, n-1} ~|~ \mu = \mu_a \right)
  $$
- Calcuting this requires the non-central t distribution.
- `power.t.test` does this very well
  - Omit one of the arguments and it solves for it

---
## Example
```{r}
power.t.test(n = 16, delta = 2 / 4, sd=1, type = "one.sample",  alt = "one.sided")$power
power.t.test(n = 16, delta = 2, sd=4, type = "one.sample",  alt = "one.sided")$power
power.t.test(n = 16, delta = 100, sd=200, type = "one.sample", alt = "one.sided")$power
```

---
## Example
```{r}
power.t.test(power = .8, delta = 2 / 4, sd=1, type = "one.sample",  alt = "one.sided")$n
power.t.test(power = .8, delta = 2, sd=4, type = "one.sample",  alt = "one.sided")$n
power.t.test(power = .8, delta = 100, sd=200, type = "one.sample", alt = "one.sided")$n
```

