---
title       : What is prediction?
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## The central dogma of prediction

<img class="center" src=../../assets/img/08_PredictionAndMachineLearning/centraldogma.png height=450>

---

## What can go wrong

<img class="center" src=../../assets/img/08_PredictionAndMachineLearning/googleflu.png height=450>


http://www.sciencemag.org/content/343/6176/1203.full.pdf

---

## Components of a predictor

</br>

<center> question -> input data -> features -> algorithm -> parameters -> evaluation  </center>


---

## SPAM Example

</br>

<center> <redtext>question</redtext> -> input data -> features -> algorithm -> parameters -> evaluation  </center>

</br>

__Start with a general question__

Can I automatically detect emails that are SPAM that are not?

__Make it concrete__

Can I use quantitative characteristics of the emails to classify them as SPAM/HAM?


---

## SPAM Example

</br>

<center> question -> <redtext>input data </redtext> -> features -> algorithm -> parameters -> evaluation  </center>

<img class=center src=../../assets/img/spamR.png height='400' />

[http://rss.acs.unt.edu/Rdoc/library/kernlab/html/spam.html](http://rss.acs.unt.edu/Rdoc/library/kernlab/html/spam.html)



---

## SPAM Example

</br>

<center> question -> input data  -> <redtext>features</redtext> -> algorithm -> parameters -> evaluation  </center>

</br>



<b>
Dear Jeff, 

Can you send me your address so I can send you the invitation? 

Thanks,

Ben
</b>




---

## SPAM Example

</br>

<center> question -> input data  -> <redtext>features</redtext> -> algorithm -> parameters -> evaluation  </center>

</br>

<b> 

Dear Jeff, 

Can <rt>you</rt> send me your address so I can send <rt>you</rt> the invitation? 

Thanks,

Ben
</b>

</br>

Frequency of you $= 2/17 = 0.118$

---

## SPAM Example

</br>

<center> question -> input data  -> <redtext>features</redtext> -> algorithm -> parameters -> evaluation  </center>


```{r loadData}
library(kernlab)
data(spam)
head(spam)

```


---

## SPAM Example


<center> question -> input data  -> features -> <redtext>algorithm</redtext> -> parameters -> evaluation  </center>

```{r,dependson="loadData",fig.height=3.5,fig.width=3.5}
plot(density(spam$your[spam$type=="nonspam"]),
     col="blue",main="",xlab="Frequency of 'your'")
lines(density(spam$your[spam$type=="spam"]),col="red")
```

---

## SPAM Example


<center> question -> input data  -> features -> <redtext>algorithm</redtext> -> parameters -> evaluation  </center>

</br></br>

__Our algorithm__

* Find a value $C$. 
* __frequency of 'your' $>$ C__ predict "spam"

---

## SPAM Example


<center> question -> input data  -> features -> algorithm -> <redtext>parameters</redtext> -> evaluation  </center>

```{r,dependson="loadData",fig.height=3.5,fig.width=3.5}
plot(density(spam$your[spam$type=="nonspam"]),
     col="blue",main="",xlab="Frequency of 'your'")
lines(density(spam$your[spam$type=="spam"]),col="red")
abline(v=0.5,col="black")
```

---

## SPAM Example


<center> question -> input data  -> features -> algorithm -> parameters -> <redtext>evaluation</redtext></center>

```{r,dependson="loadData",fig.height=3.5,fig.width=3.5}
prediction <- ifelse(spam$your > 0.5,"spam","nonspam")
table(prediction,spam$type)/length(spam$type)
```

Accuracy$ \approx 0.459 + 0.292 = 0.751$
