---
title       : In sample and out of sample error
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## In sample versus out of sample

__In Sample Error__: The error rate you get on the same
data set you used to build your predictor. Sometimes
called resubstitution error.

__Out of Sample Error__: The error rate you get on a new
data set. Sometimes called generalization error. 

__Key ideas__

1. Out of sample error is what you care about
2. In sample error $<$ out of sample error
3. The reason is overfitting
  * Matching your algorithm to the data you have

---

## In sample versus out of sample errors

```{r loadData, fig.height=3.5,fig.width=3.5}
library(kernlab); data(spam); set.seed(333)
smallSpam <- spam[sample(dim(spam)[1],size=10),]
spamLabel <- (smallSpam$type=="spam")*1 + 1
plot(smallSpam$capitalAve,col=spamLabel)
```


---

## Prediction rule 1

* capitalAve $>$ 2.7 = "spam"
* capitalAve $<$ 2.40 = "nonspam"
* capitalAve between 2.40 and 2.45 = "spam"
* capitalAve between 2.45 and 2.7 = "nonspam"

---

## Apply Rule 1 to smallSpam

```{r, dependson="loadData"}
rule1 <- function(x){
  prediction <- rep(NA,length(x))
  prediction[x > 2.7] <- "spam"
  prediction[x < 2.40] <- "nonspam"
  prediction[(x >= 2.40 & x <= 2.45)] <- "spam"
  prediction[(x > 2.45 & x <= 2.70)] <- "nonspam"
  return(prediction)
}
table(rule1(smallSpam$capitalAve),smallSpam$type)
```

---

## Prediction rule 2

* capitalAve $>$ 2.40 = "spam"
* capitalAve $\leq$ 2.40 = "nonspam"


---

## Apply Rule 2 to smallSpam


```{r, dependson="loadData"}
rule2 <- function(x){
  prediction <- rep(NA,length(x))
  prediction[x > 2.8] <- "spam"
  prediction[x <= 2.8] <- "nonspam"
  return(prediction)
}
table(rule2(smallSpam$capitalAve),smallSpam$type)
```

---

## Apply to complete spam data

```{r, dependson="loadData"}
table(rule1(spam$capitalAve),spam$type)
table(rule2(spam$capitalAve),spam$type)
mean(rule1(spam$capitalAve)==spam$type)
mean(rule2(spam$capitalAve)==spam$type)
```

---

## Look at accuracy

```{r, dependson="loadData"}
sum(rule1(spam$capitalAve)==spam$type)
sum(rule2(spam$capitalAve)==spam$type)
```


---

## What's going on? 

<center><rt> Overfitting </rt></center>

* Data have two parts
  * Signal
  * Noise
* The goal of a predictor is to find signal
* You can always design a perfect in-sample predictor
* You capture both signal + noise when you do that
* Predictor won't perform as well on new samples

[http://en.wikipedia.org/wiki/Overfitting](http://en.wikipedia.org/wiki/Overfitting)