---
title       : Reading local flat files
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Example - Baltimore camera data

<img class=center src=../../assets/img/03_ObtainingData/cameras.png height=500>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)

---

## Download the file to load

```{r data}
if(!file.exists("data")){dir.create("data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="cameras.csv",method="curl")
dateDownloaded <- date()
```

---

## Loading flat files - read.table()

* This is the main function for reading data into R
* Flexible and robust but requires more parameters
* Reads the data into RAM - big data can cause problems
* Important parameters _file_, _header_, _sep_, _row.names_, _nrows_
* Related: _read.csv()_, _read.csv2()_


---

## Baltimore example


```{r ,dependson="data"}
cameraData <- read.table("./data/cameras.csv")
head(cameraData)
```

---

## Example: Baltimore camera data

```{r}
cameraData <- read.table("./data/cameras.csv",sep=",",header=TRUE)
head(cameraData)
```

---

## Example: Baltimore camera data

read.csv sets _sep=","_ and _header=TRUE_ 
```{r,dependson="data"}
cameraData <- read.csv("./data/cameras.csv")
head(cameraData)
```


---

## Some more important parameters

* _quote_ - you can tell R whether there are any quoted values quote="" means no quotes.
* _na.strings_ - set the character that represents a missing value. 
* _nrows_ - how many rows to read of the file (e.g. nrows=10 reads 10 lines).
* _skip_ - number of lines to skip before starting to read

_In my experience, the biggest trouble with reading flat files are quotation marks ` or " placed in data values, setting quote="" often resolves these_.