---
title       : Preprocessing with Principal Components Analysis (PCA)
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Correlated predictors

```{r loadPackage,cache=TRUE,fig.height=3.5,fig.width=3.5}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]

M <- abs(cor(training[,-58]))
diag(M) <- 0
which(M > 0.8,arr.ind=T)
```

---

## Correlated predictors

```{r,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
names(spam)[c(34,32)]
plot(spam[,34],spam[,32])
```


---

## Basic PCA idea

* We might not need every predictor
* A weighted combination of predictors might be better
* We should pick this combination to capture the "most information" possible
* Benefits
  * Reduced number of predictors
  * Reduced noise (due to averaging)


---

## We could rotate the plot

$$ X = 0.71 \times {\rm num 415} + 0.71 \times {\rm num857}$$

$$ Y = 0.71 \times {\rm num 415} - 0.71 \times {\rm num857}$$

```{r,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
X <- 0.71*training$num415 + 0.71*training$num857
Y <- 0.71*training$num415 - 0.71*training$num857
plot(X,Y)
```

---

## Related problems

You have multivariate variables $X_1,\ldots,X_n$ so $X_1 = (X_{11},\ldots,X_{1m})$

* Find a new set of multivariate variables that are uncorrelated and explain as much variance as possible.
* If you put all the variables together in one matrix, find the best matrix created with fewer variables (lower rank) that explains the original data.


The first goal is <font color="#330066">statistical</font> and the second goal is <font color="#993300">data compression</font>.

---

## Related solutions - PCA/SVD

__SVD__

If $X$ is a matrix with each variable in a column and each observation in a row then the SVD is a "matrix decomposition"

$$ X = UDV^T$$

where the columns of $U$ are orthogonal (left singular vectors), the columns of $V$ are orthogonal (right singluar vectors) and $D$ is a diagonal matrix (singular values). 

__PCA__

The principal components are equal to the right singular values if you first scale (subtract the mean, divide by the standard deviation) the variables.

---

## Principal components in R - prcomp

```{r prcomp,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
smallSpam <- spam[,c(34,32)]
prComp <- prcomp(smallSpam)
plot(prComp$x[,1],prComp$x[,2])
```

---

## Principal components in R - prcomp

```{r ,dependson="prcomp",cache=TRUE,fig.height=3.5,fig.width=3.5}
prComp$rotation
```


---

## PCA on SPAM data

```{r spamPC,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
typeColor <- ((spam$type=="spam")*1 + 1)
prComp <- prcomp(log10(spam[,-58]+1))
plot(prComp$x[,1],prComp$x[,2],col=typeColor,xlab="PC1",ylab="PC2")
```


---

## PCA with caret

```{r ,dependson="spamPC",cache=TRUE,fig.height=3.5,fig.width=3.5}
preProc <- preProcess(log10(spam[,-58]+1),method="pca",pcaComp=2)
spamPC <- predict(preProc,log10(spam[,-58]+1))
plot(spamPC[,1],spamPC[,2],col=typeColor)
```


---

## Preprocessing with PCA

```{r pcaCaret,dependson="spamPC",cache=TRUE,fig.height=3.5,fig.width=3.5}
preProc <- preProcess(log10(training[,-58]+1),method="pca",pcaComp=2)
trainPC <- predict(preProc,log10(training[,-58]+1))
modelFit <- train(training$type ~ .,method="glm",data=trainPC)
```

---

## Preprocessing with PCA

```{r ,dependson="pcaCaret",cache=TRUE,fig.height=3.5,fig.width=3.5}
testPC <- predict(preProc,log10(testing[,-58]+1))
confusionMatrix(testing$type,predict(modelFit,testPC))
```

---

## Alternative (sets # of PCs)

```{r ,dependson="pcaCaret",cache=TRUE,fig.height=3.5,fig.width=3.5}
modelFit <- train(training$type ~ .,method="glm",preProcess="pca",data=training)
confusionMatrix(testing$type,predict(modelFit,testing))
```

---

## Final thoughts on PCs

* Most useful for linear-type models
* Can make it harder to interpret predictors
* Watch out for outliers! 
  * Transform first (with logs/Box Cox)
  * Plot predictors to identify problems
* For more info see 
  * Exploratory Data Analysis
  * [Elements of Statistical Learning](http://statweb.stanford.edu/~tibs/ElemStatLearn/)
  