---
title       : Principal Components Analysis and Singular Value Decomposition
subtitle    : 
author      : Roger D. Peng, Associate Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


## Matrix data 

```{r randomData,fig.height=4,fig.width=3}
set.seed(12345); par(mar=rep(0.2,4))
dataMatrix <- matrix(rnorm(400),nrow=40)
image(1:10,1:40,t(dataMatrix)[,nrow(dataMatrix):1])
```

---

## Cluster the data 

```{r,fig.height=6,fig.width=6}
par(mar=rep(0.2,4))
heatmap(dataMatrix)
```

---

## What if we add a pattern?

```{r onePattern}
set.seed(678910)
for(i in 1:40){
  # flip a coin
  coinFlip <- rbinom(1,size=1,prob=0.5)
  # if coin is heads add a common pattern to that row
  if(coinFlip){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,3),each=5)
  }
}
```


---

## What if we add a pattern? - the data

```{r dependson ="onePattern",fig.height=5,fig.width=4}
par(mar=rep(0.2,4))
image(1:10,1:40,t(dataMatrix)[,nrow(dataMatrix):1])
```


---

## What if we add a pattern? - the clustered data

```{r dependson ="onePattern",fig.height=6,fig.width=6}
par(mar=rep(0.2,4))
heatmap(dataMatrix)
```



---

## Patterns in rows and columns



```{r oChunk, dependson ="onePattern",fig.height=4,fig.width=12}
hh <- hclust(dist(dataMatrix)); dataMatrixOrdered <- dataMatrix[hh$order,]
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(rowMeans(dataMatrixOrdered),40:1,,xlab="Row Mean",ylab="Row",pch=19)
plot(colMeans(dataMatrixOrdered),xlab="Column",ylab="Column Mean",pch=19)
```

---

## Related problems

You have multivariate variables $X_1,\ldots,X_n$ so $X_1 = (X_{11},\ldots,X_{1m})$

* Find a new set of multivariate variables that are uncorrelated and explain as much variance as possible.
* If you put all the variables together in one matrix, find the best matrix created with fewer variables (lower rank) that explains the original data.


The first goal is <font color="#330066">statistical</font> and the second goal is <font color="#993300">data compression</font>.

---

## Related solutions - PCA/SVD

__SVD__

If $X$ is a matrix with each variable in a column and each observation in a row then the SVD is a "matrix decomposition"

$$ X = UDV^T$$

where the columns of $U$ are orthogonal (left singular vectors), the columns of $V$ are orthogonal (right singular vectors) and $D$ is a diagonal matrix (singular values). 

__PCA__

The principal components are equal to the right singular values if you first scale (subtract the mean, divide by the standard deviation) the variables.

---

## Components of the SVD - $u$ and $v$

```{r dependson ="oChunk",fig.height=4,fig.width=12}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(svd1$u[,1],40:1,,xlab="Row",ylab="First left singular vector",pch=19)
plot(svd1$v[,1],xlab="Column",ylab="First right singular vector",pch=19)
```


---

## Components of the SVD - Variance explained

```{r dependson ="oChunk",fig.height=5,fig.width=10}
par(mfrow=c(1,2))
plot(svd1$d,xlab="Column",ylab="Singular value",pch=19)
plot(svd1$d^2/sum(svd1$d^2),xlab="Column",ylab="Prop. of variance explained",pch=19)
```


---

## Relationship to principal components

```{r dependson ="oChunk",fig.height=5,fig.width=5}
svd1 <- svd(scale(dataMatrixOrdered))
pca1 <- prcomp(dataMatrixOrdered,scale=TRUE)
plot(pca1$rotation[,1],svd1$v[,1],pch=19,xlab="Principal Component 1",ylab="Right Singular Vector 1")
abline(c(0,1))
```

---

## Components of the SVD - variance explained

```{r dependson ="oChunk",fig.height=4,fig.width=12,tidy=FALSE}
constantMatrix <- dataMatrixOrdered*0
for(i in 1:dim(dataMatrixOrdered)[1]){constantMatrix[i,] <- rep(c(0,1),each=5)}
svd1 <- svd(constantMatrix)
par(mfrow=c(1,3))
image(t(constantMatrix)[,nrow(constantMatrix):1])
plot(svd1$d,xlab="Column",ylab="Singular value",pch=19)
plot(svd1$d^2/sum(svd1$d^2),xlab="Column",ylab="Prop. of variance explained",pch=19)
```


---

## What if we add a second pattern?

```{r twoPattern, dependson ="randomData",fig.height=4,fig.width=3}
set.seed(678910)
for(i in 1:40){
  # flip a coin
  coinFlip1 <- rbinom(1,size=1,prob=0.5)
  coinFlip2 <- rbinom(1,size=1,prob=0.5)
  # if coin is heads add a common pattern to that row
  if(coinFlip1){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5),each=5)
  }
  if(coinFlip2){
    dataMatrix[i,] <- dataMatrix[i,] + rep(c(0,5),5)
  }
}
hh <- hclust(dist(dataMatrix)); dataMatrixOrdered <- dataMatrix[hh$order,]
```

---

## Singular value decomposition - true patterns 

```{r  dependson ="twoPattern",fig.height=4.5,fig.width=12}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(rep(c(0,1),each=5),pch=19,xlab="Column",ylab="Pattern 1")
plot(rep(c(0,1),5),pch=19,xlab="Column",ylab="Pattern 2")
```

---

##  $v$ and patterns of variance in rows

```{r  dependson ="twoPattern",fig.height=4.5,fig.width=12}
svd2 <- svd(scale(dataMatrixOrdered))
par(mfrow=c(1,3))
image(t(dataMatrixOrdered)[,nrow(dataMatrixOrdered):1])
plot(svd2$v[,1],pch=19,xlab="Column",ylab="First right singular vector")
plot(svd2$v[,2],pch=19,xlab="Column",ylab="Second right singular vector")
```


---

##  $d$ and variance explained

```{r dependson ="twoPattern",fig.height=4,fig.width=8}
svd1 <- svd(scale(dataMatrixOrdered))
par(mfrow=c(1,2))
plot(svd1$d,xlab="Column",ylab="Singular value",pch=19)
plot(svd1$d^2/sum(svd1$d^2),xlab="Column",ylab="Percent of variance explained",pch=19)
```

---

## Missing values

```{r,dependson="twoPattern",error=TRUE}
dataMatrix2 <- dataMatrixOrdered
## Randomly insert some missing data
dataMatrix2[sample(1:100,size=40,replace=FALSE)] <- NA
svd1 <- svd(scale(dataMatrix2))  ## Doesn't work!
```


---

## Imputing {impute}

```{r,dependson="twoPattern",fig.height=4,fig.width=8,tidy=FALSE}
library(impute)  ## Available from http://bioconductor.org
dataMatrix2 <- dataMatrixOrdered
dataMatrix2[sample(1:100,size=40,replace=FALSE)] <- NA
dataMatrix2 <- impute.knn(dataMatrix2)$data
svd1 <- svd(scale(dataMatrixOrdered)); svd2 <- svd(scale(dataMatrix2))
par(mfrow=c(1,2)); plot(svd1$v[,1],pch=19); plot(svd2$v[,1],pch=19)
```



---

## Face example

<!-- ## source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R") -->

```{r loadFaceData ,fig.height=6,fig.width=6}
load("data/face.rda")
image(t(faceData)[,nrow(faceData):1])
```


---

## Face example - variance explained

```{r,dependson="loadFaceData",fig.height=5,fig.width=6}
svd1 <- svd(scale(faceData))
plot(svd1$d^2/sum(svd1$d^2),pch=19,xlab="Singular vector",ylab="Variance explained")
```

---

## Face example - create approximations

```{r approximations,dependson="loadFaceData",fig.height=4,fig.width=4}

svd1 <- svd(scale(faceData))
## Note that %*% is matrix multiplication

# Here svd1$d[1] is a constant
approx1 <- svd1$u[,1] %*% t(svd1$v[,1]) * svd1$d[1]

# In these examples we need to make the diagonal matrix out of d
approx5 <- svd1$u[,1:5] %*% diag(svd1$d[1:5])%*% t(svd1$v[,1:5]) 
approx10 <- svd1$u[,1:10] %*% diag(svd1$d[1:10])%*% t(svd1$v[,1:10]) 
```

---

## Face example - plot approximations
```{r dependson="approximations",fig.height=4,fig.width=14}
par(mfrow=c(1,4))
image(t(approx1)[,nrow(approx1):1], main = "(a)")
image(t(approx5)[,nrow(approx5):1], main = "(b)")
image(t(approx10)[,nrow(approx10):1], main = "(c)")
image(t(faceData)[,nrow(faceData):1], main = "(d)")  ## Original data
```


---

## Notes and further resources

* Scale matters
* PC's/SV's may mix real patterns
* Can be computationally intensive
* [Advanced data analysis from an elementary point of view](http://www.stat.cmu.edu/~cshalizi/ADAfaEPoV/ADAfaEPoV.pdf)
* [Elements of statistical learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
* Alternatives
  * [Factor analysis](http://en.wikipedia.org/wiki/Factor_analysis)
  * [Independent components analysis](http://en.wikipedia.org/wiki/Independent_component_analysis)
  * [Latent semantic analysis](http://en.wikipedia.org/wiki/Latent_semantic_analysis)









