---
title       : Basic least squares
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Goals of statistical modeling

* Describe the distribution of variables
* Describe the relationship between variables
* Make inferences about distributions or relationships


---

## Example: Average parent and child heights



<img class=center src=../../assets/img/galton.jpg height=450>

[http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html](http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html)


---


## Still relevant


<img class=center src=../../assets/img/height.png height=350>

[http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html](http://www.nature.com/ejhg/journal/v17/n8/full/ejhg20095a.html)

[Predicting height: the Victorian approach beats modern genomics](http://www.wired.com/wiredscience/2009/03/predicting-height-the-victorian-approach-beats-modern-genomics/)


---

## Load Galton Data

You may need to run `install.packages("UsingR")` if the `UsingR` library is not installed

```{r galton,fig.height=3.5,fig.width=8}
library(UsingR); data(galton)
par(mfrow=c(1,2))
hist(galton$child,col="blue",breaks=100)
hist(galton$parent,col="blue",breaks=100)
```

---

## The distribution of child heights

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
```


---

## Only know the child  - average height

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
meanChild <- mean(galton$child)
lines(rep(meanChild,100),seq(0,150,length=100),col="red",lwd=5)
```

---

## Only know the child  - why average?

If $C_i$ is the height of child $i$ then the average is the value of $\mu$ that minimizes:

$$ \sum_{i=1}^{928}(C_i - \mu)^2 $$

---

## What if we plot child versus average parent

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
```

---

## Jittered plot

```{r, dependson="galton",fig.height=4,fig.width=4}
set.seed(1234)
plot(jitter(galton$parent,factor=2),jitter(galton$child,factor=2),pch=19,col="blue")
```

---

## Average parent = 65 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near65 <- galton[abs(galton$parent - 65)<1, ]
points(near65$parent,near65$child,pch=19,col="red")
lines(seq(64,66,length=100),rep(mean(near65$child),100),col="red",lwd=4)
```

---

## Average parent = 71 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near71 <- galton[abs(galton$parent - 71)<1, ]
points(near71$parent,near71$child,pch=19,col="red")
lines(seq(70,72,length=100),rep(mean(near71$child),100),col="red",lwd=4)
```


---

## Fitting a line



```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lm1 <- lm(galton$child ~ galton$parent)
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

---

## Why not this line?


```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent, 26 + 0.646*galton$parent)
```

---

## The equation for a line

If $C_i$ is the height of child $i$ and $P_i$ is the height of the average parent, then we can imagine writing the equation for a line

$$C_i = b_0 + b_1 P_i$$


---

## Not all points are on the line

```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```


---

## Allowing for variation

If $C_i$ is the height of child $i$ and $P_i$ is the height of the average parent, then we can imagine writing the equation for a line

$$C_i = b_0 + b_1 P_i + e_i$$

$e_i$ is everything we didn't measure (how much they eat, where they live, do they stretch in the morning...)

---

## How do we pick best?

If $C_i$ is the height of child $i$ and $P_i$ is the height of the average parent, pick the line that makes the child values $C_i$ and our guesses

$$ \sum_{i=1}^{928}(C_i - \{b_0 + b_1 P_i\})^2 $$


---

## Plot what is leftover

```{r, dependson="lm1",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
plot(galton$parent,lm1$residuals,col="blue",pch=19)
abline(c(0,0),col="red",lwd=3)
```

