---
title       : Residuals and residual variation
subtitle    : 
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}

---
```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```
## Residuals
* Model $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$ where $\epsilon_i \sim N(0, \sigma^2)$.
* Observed outcome $i$ is $Y_i$ at predictor value $X_i$
* Predicted outcome $i$ is $\hat Y_i$ at predictor valuve $X_i$ is
  $$
  \hat Y_i = \hat \beta_0 + \hat \beta_1 X_i
  $$
* Residual, the between the observed and predicted outcome
  $$
  e_i = Y_i - \hat Y_i
  $$
  * The vertical distance between the observed data point and the regression line
* Least squares minimizes $\sum_{i=1}^n e_i^2$
* The $e_i$ can be thought of as estimates of the $\epsilon_i$.

---
## Properties of the residuals
* $E[e_i] = 0$.
* If an intercept is included, $\sum_{i=1}^n e_i = 0$
* If a regressor variable, $X_i$, is included in the model $\sum_{i=1}^n e_i X_i = 0$. 
* Residuals are useful for investigating poor model fit.
* Positive residuals are above the line, negative residuals are below.
* Residuals can be thought of as the outcome ($Y$) with the
  linear association of the predictor ($X$) removed.
* One differentiates residual variation (variation after removing
the predictor) from systematic variation (variation explained by the regression model).
* Residual plots highlight poor model fit.

---
## Code

```{r}
data(diamond)
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
e <- resid(fit)
yhat <- predict(fit)
max(abs(e -(y - yhat)))
max(abs(e - (y - coef(fit)[1] - coef(fit)[2] * x)))
```

---
## Residuals are the signed length of the red lines
```{r, echo = FALSE, fig.height=5, fig.width=5}
plot(diamond$carat, diamond$price,  
     xlab = "Mass (carats)", 
     ylab = "Price (SIN $)", 
     bg = "lightblue", 
     col = "black", cex = 1.1, pch = 21,frame = FALSE)
abline(fit, lwd = 2)
for (i in 1 : n) 
  lines(c(x[i], x[i]), c(y[i], yhat[i]), col = "red" , lwd = 2)
```

---
## Residuals versus X
```{r, echo = FALSE, fig.height=5, fig.width=5}
plot(diamond$carat, e,  
     xlab = "Mass (carats)", 
     ylab = "Residuals (SIN $)", 
     bg = "lightblue", 
     col = "black", cex = 1.1, pch = 21,frame = FALSE)
abline(h = 0, lwd = 2)
for (i in 1 : n) 
  lines(c(x[i], x[i]), c(e[i], 0), col = "red" , lwd = 2)
```

---
## Non-linear data
```{r, echo = TRUE, fig.height=5, fig.width=5}
x <- runif(100, -3, 3); y <- x + sin(x) + rnorm(100, sd = .2); 
plot(x, y); abline(lm(y ~ x))
```

---
```{r, echo = TRUE, fig.height=5, fig.width=5}
plot(x, resid(lm(y ~ x))); 
abline(h = 0)
```

---
## Heteroskedasticity
```{r, echo = TRUE, fig.height=4.5, fig.width=4.5}
x <- runif(100, 0, 6); y <- x + rnorm(100,  mean = 0, sd = .001 * x); 
plot(x, y); abline(lm(y ~ x))
```

---
## Getting rid of the blank space can be helpful
```{r, echo = TRUE, fig.height=4.5, fig.width=4.5}
plot(x, resid(lm(y ~ x))); 
abline(h = 0)
```

---
## Estimating residual variation
* Model $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$ where $\epsilon_i \sim N(0, \sigma^2)$.
* The ML estimate of $\sigma^2$ is $\frac{1}{n}\sum_{i=1}^n e_i^2$,
the average squared residual. 
* Most people use
  $$
  \hat \sigma^2 = \frac{1}{n-2}\sum_{i=1}^n e_i^2.
  $$
* The $n-2$ instead of $n$ is so that $E[\hat \sigma^2] = \sigma^2$

---
## Diamond example
```{r, echo = TRUE}
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
summary(fit)$sigma
sqrt(sum(resid(fit)^2) / (n - 2))
```

---
## Summarizing variation
$$
\begin{align}
\sum_{i=1}^n (Y_i - \bar Y)^2 
& = \sum_{i=1}^n (Y_i - \hat Y_i + \hat Y_i - \bar Y)^2 \\
& = \sum_{i=1}^n (Y_i - \hat Y_i)^2 + 
2 \sum_{i=1}^n  (Y_i - \hat Y_i)(\hat Y_i - \bar Y) + 
\sum_{i=1}^n  (\hat Y_i - \bar Y)^2 \\
\end{align}
$$

****
### Scratch work
$(Y_i - \hat Y_i) = \{Y_i - (\bar Y - \hat \beta_1 \bar X) - \hat \beta_1 X_i\} = (Y_i - \bar Y) - \hat \beta_1 (X_i - \bar X)$

$(\hat Y_i - \bar Y) = (\bar Y - \hat \beta_1 \bar X - \hat \beta_1 X_i - \bar Y )
= \hat \beta_1  (X_i - \bar X)$

$\sum_{i=1}^n  (Y_i - \hat Y_i)(\hat Y_i - \bar Y) 
= \sum_{i=1}^n  \{(Y_i - \bar Y) - \hat \beta_1 (X_i - \bar X))\}\{\hat \beta_1  (X_i - \bar X)\}$

$=\hat \beta_1 \sum_{i=1}^n (Y_i - \bar Y)(X_i - \bar X) -\hat\beta_1^2\sum_{i=1}^n (X_i - \bar X)^2$

$= \hat \beta_1^2 \sum_{i=1}^n (X_i - \bar X)^2-\hat\beta_1^2\sum_{i=1}^n (X_i - \bar X)^2 = 0$

---
## Summarizing variation
$$
\sum_{i=1}^n (Y_i - \bar Y)^2 
= \sum_{i=1}^n (Y_i - \hat Y_i)^2 + \sum_{i=1}^n  (\hat Y_i - \bar Y)^2 
$$

Or 

Total Variation = Residual Variation + Regression Variation

Define the percent of total varation described by the model as
$$
R^2 = \frac{\sum_{i=1}^n  (\hat Y_i - \bar Y)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2}
= 1 - \frac{\sum_{i=1}^n  (Y_i - \hat Y_i)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2}
$$

---
## Relation between $R^2$ and $r$ (the corrrelation)
Recall that $(\hat Y_i - \bar Y) = \hat \beta_1  (X_i - \bar X)$
so that
$$
R^2 = \frac{\sum_{i=1}^n  (\hat Y_i - \bar Y)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2}
= \hat \beta_1^2  \frac{\sum_{i=1}^n(X_i - \bar X)^2}{\sum_{i=1}^n (Y_i - \bar Y)^2}
= Cor(Y, X)^2
$$
Since, recall, 
$$
\hat \beta_1 = Cor(Y, X)\frac{Sd(Y)}{Sd(X)}
$$
So, $R^2$ is literally $r$ squared.

---
## Some facts about $R^2$
* $R^2$ is the percentage of variation explained by the regression model.
* $0 \leq R^2 \leq 1$
* $R^2$ is the sample correlation squared.
* $R^2$ can be a misleading summary of model fit. 
  * Deleting data can inflate $R^2$.
  * (For later.) Adding terms to a regression model always increases $R^2$.
* Do `example(anscombe)` to see the following data.
  * Basically same mean and variance of X and Y.
  * Identical correlations (hence same $R^2$ ).
  * Same linear regression relationship.

---
## `data(anscombe);example(anscombe)`
```{r, echo = FALSE, fig.height=5, fig.width=5, results='hide'}
require(stats); require(graphics); data(anscombe)
ff <- y ~ x
mods <- setNames(as.list(1:4), paste0("lm", 1:4))
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  ## or   ff[[2]] <- as.name(paste0("y", i))
  ##      ff[[3]] <- as.name(paste0("x", i))
  mods[[i]] <- lmi <- lm(ff, data = anscombe)
  #print(anova(lmi))
}


## Now, do what you should have done in the first place: PLOTS
op <- par(mfrow = c(2, 2), mar = 0.1+c(4,4,1,1), oma =  c(0, 0, 2, 0))
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  plot(ff, data = anscombe, col = "red", pch = 21, bg = "orange", cex = 1.2,
       xlim = c(3, 19), ylim = c(3, 13))
  abline(mods[[i]], col = "blue")
}
mtext("Anscombe's 4 Regression data sets", outer = TRUE, cex = 1.5)
par(op)
```

