---
title       : Using data.table
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=FALSE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## data.table

* Inherets from data.frame
  * All functions that accept data.frame work on data.table
* Written in C so it is much faster
* Much, much faster at subsetting, group, and updating
 

---

## Create data tables just like data frames

```{r init}
library(data.table)
DF = data.frame(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9))
head(DF,3)
DT = data.table(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9))
head(DT,3)
```

---

## See all the data tables in memory

```{r,dependson="init"}
tables()
```

---

## Subsetting rows

```{r,dependson="init"}
DT[2,]
DT[DT$y=="a",]
```

---

## Subsetting rows

```{r,dependson="init"}
DT[c(2,3)]
```


---

## Subsetting columns!?

```{r,dependson="init"}
DT[,c(2,3)]
```

---

## Column subsetting in data.table

* The subsetting function is modified for data.table
* The argument you pass after the comma is called an "expression"
* In R an expression is a collection of statements enclosed in curley brackets 
```{r }
{
  x = 1
  y = 2
}
k = {print(10); 5}
print(k)
```

---

## Calculating values for variables with expressions

```{r,dependson="init"}
DT[,list(mean(x),sum(z))]
DT[,table(y)]
```

---

## Adding new columns

```{r,dependson="init"}
DT[,w:=z^2]
```


---

## Adding new columns

```{r dt2,dependson="init"}
DT2 <- DT
DT[, y:= 2]
```

---

## Careful

```{r ,dependson="dt2"}
head(DT,n=3)
head(DT2,n=3)
```


---

## Multiple operations

```{r,dependson="init"}
DT[,m:= {tmp <- (x+z); log2(tmp+5)}]
```

---

## plyr like operations

```{r,dependson="init"}
DT[,a:=x>0]
```


---

## plyr like operations

```{r,dependson="init"}
DT[,b:= mean(x+w),by=a]
```


---

## Special variables

`.N` An integer, length 1, containing the number of elements of a factor level 

```{r}
set.seed(123);
DT <- data.table(x=sample(letters[1:3], 1E5, TRUE))
DT[, .N, by=x]
```

---

## Keys

```{r}
DT <- data.table(x=rep(c("a","b","c"),each=100), y=rnorm(300))
setkey(DT, x)
DT['a']
```

---

## Joins

```{r}
DT1 <- data.table(x=c('a', 'a', 'b', 'dt1'), y=1:4)
DT2 <- data.table(x=c('a', 'b', 'dt2'), z=5:7)
setkey(DT1, x); setkey(DT2, x)
merge(DT1, DT2)
```



---

## Fast reading

```{r,cache=TRUE}
big_df <- data.frame(x=rnorm(1E6), y=rnorm(1E6))
file <- tempfile()
write.table(big_df, file=file, row.names=FALSE, col.names=TRUE, sep="\t", quote=FALSE)
system.time(fread(file))
system.time(read.table(file, header=TRUE, sep="\t"))
```




---

## Summary and further reading

* The latest development version contains new functions like `melt` and `dcast` for data.tables 
  * [https://r-forge.r-project.org/scm/viewvc.php/pkg/NEWS?view=markup&root=datatable](https://r-forge.r-project.org/scm/viewvc.php/pkg/NEWS?view=markup&root=datatable)
* Here is a list of differences between data.table and data.frame
  * [http://stackoverflow.com/questions/13618488/what-you-can-do-with-data-frame-that-you-cant-in-data-table](http://stackoverflow.com/questions/13618488/what-you-can-do-with-data-frame-that-you-cant-in-data-table)
* Notes based on Raphael Gottardo's notes [https://github.com/raphg/Biostat-578/blob/master/Advanced_data_manipulation.Rpres](https://github.com/raphg/Biostat-578/blob/master/Advanced_data_manipulation.Rpres), who got them from Kevin Ushey.