---
title       : The ggplot2 Plotting System - Part 1
subtitle    : 
author      : Roger D. Peng, Associate Professor of Biostatistics
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 80)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', fig.height = 4, cache = TRUE, fig.show = 'hold')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## What is ggplot2?

- An implementation of _The Grammar of Graphics_ by Leland Wilkinson
- Written by Hadley Wickham (while he was a graduate student at Iowa State)
- A “third” graphics system for R (along with __base__ and __lattice__)
- Available from CRAN via `install.packages()`
- Web site: http://ggplot2.org (better documentation)

---

## What is ggplot2?

- Grammar of graphics represents an abstraction of graphics ideas/objects
- Think “verb”, “noun”, “adjective” for graphics
- Allows for a “theory” of graphics on which to build new graphics and graphics objects
- “Shorten the distance from mind to page”

---

## Grammer of Graphics

### “In brief, the grammar tells us that a statistical graphic is a __mapping__ from data to __aesthetic__ attributes (colour, shape, size) of __geometric__ objects (points, lines, bars). The plot may also contain statistical transformations of the data and is drawn on a specific coordinate system”

- from _ggplot2_ book

---

## Plotting Systems in R: Base

- “Artist’s palette” model
- Start with blank canvas and build up from there
- Start with `plot` function (or similar)
- Use annotation functions to add/modify (`text`, `lines`, `points`, `axis`)

---

## Plotting Systems in R: Base

- Convenient, mirrors how we think of building plots and analyzing data
- Can’t go back once plot has started (i.e. to adjust margins); need to plan in advance
- Difficult to “translate” to others once a new plot has been created (no graphical “language”)
  - Plot is just a series of R commands

---

## Plotting Systems in R: Lattice

- Plots are created with a single function call (`xyplot`, `bwplot`, etc.)
- Most useful for conditioning types of plots: Looking at how $y$ changes with $x$ across levels of $z$
- Things like margins/spacing set automatically because entire plot is specified at once
- Good for putting many many plots on a screen

---

## Plotting Systems in R: Lattice

- Sometimes awkward to specify an entire plot in a single function call
- Annotation in plot is not intuitive
- Use of panel functions and subscripts difficult to wield and requires intense preparation
- Cannot “add” to the plot once it’s created

---

## Plotting Systems in R: ggplot2

- Split the difference between base and lattice
- Automatically deals with spacings, text, titles but also allows you to annotate by “adding”
- Superficial similarity to lattice but generally easier/more intuitive to use
- Default mode makes many choices for you (but you _can_ customize!)

---

## The Basics: `qplot()`

- Works much like the `plot` function in base graphics system
- Looks for data in a data frame, similar to lattice, or in the parent environment
- Plots are made up of _aesthetics_ (size, shape, color) and _geoms_ (points, lines)

---

## The Basics: `qplot()`

- Factors are important for indicating subsets of the data (if they are to have different properties); they should be __labeled__
- The `qplot()` hides what goes on underneath, which is okay for most operations
- `ggplot()` is the core function and very flexible for doing things `qplot()` cannot do

---

## Example Dataset

```{r}
library(ggplot2)
str(mpg)
```


---

## ggplot2 “Hello, world!”

```{r}
qplot(displ, hwy, data = mpg)
```

---

## Modifying aesthetics

```{r}
qplot(displ, hwy, data = mpg, color = drv)

```


---

## Adding a geom

```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))

```

---

## Histograms

```{r}
qplot(hwy, data = mpg, fill = drv)

```



---

## Facets

```{r, fig.width=4.5}
qplot(displ, hwy, data = mpg, facets = . ~ drv)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

---

## MAACS Cohort

- Mouse Allergen and Asthma Cohort Study
- Baltimore children (aged 5—17)
- Persistent asthma, exacerbation in past year
- Study indoor environment and its relationship with asthma morbidity
- Recent publication: http://goo.gl/WqE9j8

```{r,echo=FALSE}
eno <- read.csv("eno.csv")
skin <- read.csv("skin.csv")
env <- read.csv("environmental.csv")
m <- merge(eno, env, by = "id")
maacs <- merge(m, skin, by = "id")
```

---

## Example: MAACS

```{r}
str(maacs)
```


---

## Histogram of eNO

```{r}
qplot(log(eno), data = maacs)
```

---

## Histogram by Group

```{r}
qplot(log(eno), data = maacs, fill = mopos)
```

---

## Density Smooth

```{r, fig.width=4.5}
qplot(log(eno), data = maacs, geom = "density")
qplot(log(eno), data = maacs, geom = "density", color = mopos)
```

---

## Scatterplots: eNO vs. PM$_{2.5}$

```{r, fig.width=2.7}
qplot(log(pm25), log(eno), data = maacs)
qplot(log(pm25), log(eno), data = maacs, shape = mopos)
qplot(log(pm25), log(eno), data = maacs, color = mopos)
```


---

## Scatterplots: eNO vs. PM$_{2.5}$

```{r}
qplot(log(pm25), log(eno), data = maacs, color = mopos, 
      geom = c("point", "smooth"), method = "lm")
```


---

## Scatterplots: eNO vs. PM$_{2.5}$

```{r, fig.width=9}
qplot(log(pm25), log(eno), data = maacs, geom = c("point", "smooth"), 
      method = "lm", facets = . ~ mopos)
```


---

## Summary of qplot()

- The `qplot()` function is the analog to `plot()` but with many built-in features
- Syntax somewhere in between base/lattice
- Produces very nice graphics, essentially publication ready (if you like the design)
- Difficult to go against the grain/customize (don’t bother; use full ggplot2 power in that case)

---

## Resources

- The _ggplot2_ book by Hadley Wickham
- The _R Graphics Cookbook_ by Winston Chang (examples in base plots and in ggplot2)
- ggplot2 web site (http://ggplot2.org)
- ggplot2 mailing list (http://goo.gl/OdW3uB), primarily for developers
