---
title       : Reading mySQL
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=TRUE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## mySQL

* Free and widely used open source database software
* Widely used in internet based applications
* Data are structured in 
  * Databases
  * Tables within databases
  * Fields within tables
* Each row is called a record

[http://en.wikipedia.org/wiki/MySQL](http://en.wikipedia.org/wiki/MySQL)
[http://www.mysql.com/](http://www.mysql.com/)


---

## Example structure

<img class=center src=../../assets/img/03_ObtainingData/database-schema.png height=450>


[http://dev.mysql.com/doc/employee/en/sakila-structure.html](http://dev.mysql.com/doc/employee/en/sakila-structure.html)

---

## Step 1 - Install MySQL

<img class=center src=../../assets/img/03_ObtainingData/installmysql.png height=450>

[http://dev.mysql.com/doc/refman/5.7/en/installing.html](http://dev.mysql.com/doc/refman/5.7/en/installing.html)

---

## Step 2 - Install RMySQL

* On a Mac: ```install.packages("RMySQL")```
* On Windows: 
  * Official instructions - [http://biostat.mc.vanderbilt.edu/wiki/Main/RMySQL](http://biostat.mc.vanderbilt.edu/wiki/Main/RMySQL) (may be useful for Mac/UNIX users as well)
  * Potentially useful guide - [http://www.ahschulz.de/2013/07/23/installing-rmysql-under-windows/](http://www.ahschulz.de/2013/07/23/installing-rmysql-under-windows/)  


---

## Example - UCSC database


<img class=center src=../../assets/img/03_ObtainingData/ucsc.png height=450>

[http://genome.ucsc.edu/](http://genome.ucsc.edu/)



---

## UCSC MySQL


<img class=center src=../../assets/img/03_ObtainingData/ucscmysql.png height=450>

[http://genome.ucsc.edu/goldenPath/help/mysql.html](http://genome.ucsc.edu/goldenPath/help/mysql.html)


---

## Connecting and listing databases

```{r databases}
ucscDb <- dbConnect(MySQL(),user="genome", 
                    host="genome-mysql.cse.ucsc.edu")
result <- dbGetQuery(ucscDb,"show databases;"); dbDisconnect(ucscDb);
result
```


---

## Connecting to hg19 and listing tables

```{r tables}
hg19 <- dbConnect(MySQL(),user="genome", db="hg19",
                    host="genome-mysql.cse.ucsc.edu")
allTables <- dbListTables(hg19)
length(allTables)
allTables[1:5]
```


---

## Get dimensions of a specific table

```{r dimensions,dependson="tables"}
dbListFields(hg19,"affyU133Plus2")
dbGetQuery(hg19, "select count(*) from affyU133Plus2")
```


---

## Read from the table

```{r readdata ,dependson="tables"}
affyData <- dbReadTable(hg19, "affyU133Plus2")
head(affyData)
```


---

## Select a specific subset

```{r, dependson="tables"}
query <- dbSendQuery(hg19, "select * from affyU133Plus2 where misMatches between 1 and 3")
affyMis <- fetch(query); quantile(affyMis$misMatches)
affyMisSmall <- fetch(query,n=10); dbClearResult(query);
dim(affyMisSmall)
```

---

## Don't forget to close the connection!

```{r, dependson="tables"}
dbDisconnect(hg19)
```

---

## Further resources

* RMySQL vignette [http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf](http://cran.r-project.org/web/packages/RMySQL/RMySQL.pdf)
* List of commands [http://www.pantz.org/software/mysql/mysqlcommands.html](http://www.pantz.org/software/mysql/mysqlcommands.html)
  * __Do not, do not, delete, add or join things from ensembl. Only select.__
  * In general be careful with mysql commands
* A nice blog post summarizing some other commands [http://www.r-bloggers.com/mysql-and-r/](http://www.r-bloggers.com/mysql-and-r/)
  