---
title       : Least squares estimation of regression lines
subtitle    : Regression via least squares
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}

---

```{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## General least squares for linear equations
Consider again the parent and child height data from Galton

```{r, fig.height=5, fig.width=5, echo=FALSE}
library(UsingR)
data(galton)
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
plot(as.numeric(as.vector(freqData$parent)), 
     as.numeric(as.vector(freqData$child)),
     pch = 21, col = "black", bg = "lightblue",
     cex = .05 * freqData$freq, 
     xlab = "parent", ylab = "child")
```

---
## Fitting the best line
* Let $Y_i$ be the $i^{th}$ child's height and $X_i$ be the 
$i^{th}$ (average over the pair of) parents' heights. 
* Consider finding the best line 
  * Child's Height = $\beta_0$ + Parent's Height $\beta_1$
* Use least squares
  $$
  \sum_{i=1}^n \{Y_i - (\beta_0 + \beta_1 X_i)\}^2
  $$
* How do we do it?

---
## Let's solve this problem generally
* Let $\mu_i = \beta_0 + \beta_1 X_i$ and our estimates be
$\hat \mu_i = \hat \beta_0 + \hat \beta_1 X_i$. 
* We want to minimize
$$ \dagger \sum_{i=1}^n (Y_i - \mu_i)^2 = \sum_{i=1}^n (Y_i - \hat \mu_i) ^ 2 + 2 \sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) + \sum_{i=1}^n (\hat \mu_i - \mu_i)^2$$
* Suppose that $$\sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) = 0$$ then
$$ \dagger 
=\sum_{i=1}^n (Y_i - \hat \mu_i) ^ 2  + \sum_{i=1}^n (\hat \mu_i - \mu_i)^2\geq \sum_{i=1}^n (Y_i - \hat \mu_i) ^ 2$$

---
## Mean only regression
* So we know that if:
$$ \sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) = 0$$
where $\mu_i = \beta_0 + \beta_1 X_i$ and $\hat \mu_i = \hat \beta_0 + \hat \beta_1 X_i$ then the line 
$$Y = \hat \beta_0 + \hat \beta_1 X$$
is the least squares line.
* Consider forcing $\beta_1 = 0$ and thus $\hat \beta_1=0$; 
that is, only considering horizontal lines
* The solution works out to be
$$\hat \beta_0 = \bar Y.$$

---
## Let's show it
$$\begin{align} \
\sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) 
= & \sum_{i=1}^n (Y_i - \hat \beta_0) (\hat \beta_0 - \beta_0) \\
= & (\hat \beta_0 - \beta_0) \sum_{i=1}^n (Y_i   - \hat \beta_0) \
\end{align} $$

Thus, this will equal 0 if $\sum_{i=1}^n (Y_i  - \hat \beta_0)
= n\bar Y - n \hat \beta_0=0$

Thus $\hat \beta_0 = \bar Y.$

---
## Regression through the origin
* Recall that if:
$$ \sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) = 0$$
where $\mu_i = \beta_0 + \beta_1 X_i$ and $\hat \mu_i = \hat \beta_0 + \hat \beta_1 X_i$ then the line 
$$Y = \hat \beta_0 + \hat \beta_1 X$$
is the least squares line.
* Consider forcing $\beta_0 = 0$ and thus $\hat \beta_0=0$; 
that is, only considering lines through the origin
* The solution works out to be
$$\hat \beta_1 = \frac{\sum_{i=1^n} Y_i X_i}{\sum_{i=1}^n X_i^2}.$$

---
## Let's show it
$$\begin{align} \
\sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) 
= & \sum_{i=1}^n (Y_i - \hat \beta_1 X_i) (\hat \beta_1 X_i - \beta_1 X_i) \\
= & (\hat \beta_1 - \beta_1) \sum_{i=1}^n (Y_i X_i  - \hat \beta_1 X_i ^2) \
\end{align} $$

Thus, this will equal 0 if $\sum_{i=1}^n (Y_i X_i  - \hat \beta_1 X_i ^2) = \sum_{i=1}^n Y_i X_i - \hat \beta_1 \sum_{i=1}^n X_i^2 =0$

Thus
$$\hat \beta_1 = \frac{\sum_{i=1^n} Y_i X_i}{\sum_{i=1}^n X_i^2}.$$


---
## Recapping what we know
* If we define $\mu_i = \beta_0$ then $\hat \beta_0 = \bar Y$.
  * If we only look at horizontal lines, the least squares estimate of the intercept of that line is the average of the outcomes.
* If we define $\mu_i = X_i \beta_1$ then $\hat \beta_1 = \frac{\sum_{i=1^n} Y_i X_i}{\sum_{i=1}^n X_i^2}$
  * If we only look at lines through the origin, we get the estimated slope is the cross product of the X and Ys divided by the cross product of the Xs with themselves.
* What about when $\mu_i = \beta_0 + \beta_1 X_i$? That is, we don't want to restrict ourselves to horizontal lines or lines through the origin.

---
## Let's figure it out
$$\begin{align} \
\sum_{i=1}^n (Y_i - \hat \mu_i) (\hat \mu_i - \mu_i) 
= & \sum_{i=1}^n (Y_i - \hat\beta_0 - \hat\beta_1 X_i) (\hat \beta_0 + \hat \beta_1 X_i - \beta_0 - \beta_1 X_i) \\
= & (\hat \beta_0 - \beta_0) \sum_{i=1}^n (Y_i - \hat\beta_0 - \hat \beta_1 X_i) + (\beta_1 - \beta_1)\sum_{i=1}^n (Y_i - \hat\beta_0 - \hat \beta_1 X_i)X_i\\
\end{align} $$
Note that 

$$0=\sum_{i=1}^n (Y_i - \hat\beta_0 - \hat \beta_1 X_i) = n \bar Y - n \hat \beta_0 - n \hat \beta_1 \bar X ~~\mbox{implies that}~~\hat \beta_0 = \bar Y - \hat \beta_1 \bar X $$

Then
$$\sum_{i=1}^n (Y_i  - \hat\beta_0 - \hat \beta_1 X_i) X_i =  \sum_{i=1}^n (Y_i  - \bar Y + \hat \beta_1 \bar X - \hat \beta_1 X_i)X_i$$

---
## Continued
$$=\sum_{i=1}^n \{(Y_i  - \bar Y) - \hat \beta_1 (X_i - \bar X) \}X_i$$
And thus
$$ \sum_{i=1}^n (Y_i  - \bar Y)X_i - \hat \beta_1 \sum_{i=1}^n
(X_i - \bar X) X_i = 0.$$
So we arrive at
$$
\hat \beta_1 =
\frac{\sum_{i=1}^n \{(Y_i  - \bar Y)X_i}{\sum_{i=1}^n
(X_i - \bar X) X_i} = 
\frac{\sum_{i=1}^n (Y_i  - \bar Y)(X_i - \bar X)}{\sum_{i=1}^n
(X_i - \bar X) (X_i - \bar X)}
= Cor(Y, X) \frac{Sd(Y)}{Sd(X)}.
$$
And recall
$$
\hat \beta_0 = \bar Y - \hat \beta_1 \bar X.
$$

---
## Consequences
* The least squares model fit to the line $Y = \beta_0 + \beta_1 X$ through the data pairs $(X_i, Y_i)$ with $Y_i$ as the outcome obtains the line $Y = \hat \beta_0 + \hat \beta_1 X$ where
  $$\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)} ~~~ \hat \beta_0 = \bar Y - \hat \beta_1 \bar X$$
* $\hat \beta_1$ has the units of $Y / X$, $\hat \beta_0$ has the units of $Y$.
* The line passes through the point $(\bar X, \bar Y$)
* The slope of the regression line with $X$ as the outcome and $Y$ as the predictor is $Cor(Y, X) Sd(X)/ Sd(Y)$. 
* The slope is the same one you would get if you centered the data,
$(X_i - \bar X, Y_i - \bar Y)$, and did regression through the origin.
* If you normalized the data, $\{ \frac{X_i - \bar X}{Sd(X)}, \frac{Y_i - \bar Y}{Sd(Y)}\}$, the slope is $Cor(Y, X)$.

---
## Revisiting Galton's data
### Double check our calculations using R
```{r, fig.height=4,fig.width=4,echo=TRUE}
y <- galton$child
x <- galton$parent
beta1 <- cor(y, x) *  sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
rbind(c(beta0, beta1), coef(lm(y ~ x)))
```

---
## Revisiting Galton's data
### Reversing the outcome/predictor relationship
```{r, fig.height=4,fig.width=4,echo=TRUE}
beta1 <- cor(y, x) *  sd(x) / sd(y)
beta0 <- mean(x) - beta1 * mean(y)
rbind(c(beta0, beta1), coef(lm(x ~ y)))
```

---
## Revisiting Galton's data
### Regression through the origin yields an equivalent slope if you center the data first
```{r, fig.height=4,fig.width=4,echo=TRUE}
yc <- y - mean(y)
xc <- x - mean(x)
beta1 <- sum(yc * xc) / sum(xc ^ 2)
c(beta1, coef(lm(y ~ x))[2])
```

---
## Revisiting Galton's data
### Normalizing variables results in the slope being the correlation
```{r, echo=TRUE}
yn <- (y - mean(y))/sd(y)
xn <- (x - mean(x))/sd(x)
c(cor(y, x), cor(yn, xn), coef(lm(yn ~ xn))[2])
```


---
## Plotting the fit
* Size of points are frequencies at that X, Y combination.
* For the red lie the child is outcome.
* For the blue, the parent is the outcome  (accounting for the fact that the response is plotted on the horizontal axis).
* Black line assumes $Cor(Y, X) = 1$ (slope is $Sd(Y)/Sd(x)$).
* Big black dot is $(\bar X, \bar Y)$.

---
The code to add the lines

```
abline(mean(y) - mean(x) * cor(y, x) * sd(y) / sd(x), 
  sd(y) / sd(x) * cor(y, x), 
  lwd = 3, col = "red")
abline(mean(y) - mean(x) * sd(y) / sd(x) / cor(y, x), 
  sd(y) cor(y, x) / sd(x), 
  lwd = 3, col = "blue")
abline(mean(y) - mean(x) * sd(y) / sd(x), 
  sd(y) / sd(x), 
  lwd = 2)
points(mean(x), mean(y), cex = 2, pch = 19)
```

---
```{r, fig.height=6,fig.width=6,echo=FALSE}
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
plot(as.numeric(as.vector(freqData$parent)), 
     as.numeric(as.vector(freqData$child)),
     pch = 21, col = "black", bg = "lightblue",
     cex = .05 * freqData$freq, 
     xlab = "parent", ylab = "child", xlim = c(62, 74), ylim = c(62, 74))
abline(mean(y) - mean(x) * cor(y, x) * sd(y) / sd(x), sd(y) / sd(x) * cor(y, x), lwd = 3, col = "red")
abline(mean(y) - mean(x) * sd(y) / sd(x) / cor(y, x), sd(y) / sd(x) / cor(y, x), lwd = 3, col = "blue")
abline(mean(y) - mean(x) * sd(y) / sd(x), sd(y) / sd(x), lwd = 2)
points(mean(x), mean(y), cex = 2, pch = 19)
```
