---
title       : Least squares estimation of regression lines
subtitle    : Regression via least squares
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}

---
````{r setup, cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE, results='hide', error=FALSE}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
runif(1)
```

## General least squares for linear equations
Consider again the parent and child height data from Galton

```{r, fig.height=5, fig.width=8, echo=FALSE}
library(UsingR)
data(galton)
library(dplyr); library(ggplot2)
freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")  
g
```

---
## Fitting the best line
* Let $Y_i$ be the $i^{th}$ child's height and $X_i$ be the 
$i^{th}$ (average over the pair of) parents' heights. 
* Consider finding the best line 
  * Child's Height = $\beta_0$ + Parent's Height $\beta_1$
* Use least squares
  $$
  \sum_{i=1}^n \{Y_i - (\beta_0 + \beta_1 X_i)\}^2
  $$

---
## Results
* The least squares model fit to the line $Y = \beta_0 + \beta_1 X$ through the data pairs $(X_i, Y_i)$ with $Y_i$ as the outcome obtains the line $Y = \hat \beta_0 + \hat \beta_1 X$ where
  $$\hat \beta_1 = Cor(Y, X) \frac{Sd(Y)}{Sd(X)} ~~~ \hat \beta_0 = \bar Y - \hat \beta_1 \bar X$$
* $\hat \beta_1$ has the units of $Y / X$, $\hat \beta_0$ has the units of $Y$.
* The line passes through the point $(\bar X, \bar Y$)
* The slope of the regression line with $X$ as the outcome and $Y$ as the predictor is $Cor(Y, X) Sd(X)/ Sd(Y)$. 
* The slope is the same one you would get if you centered the data,
$(X_i - \bar X, Y_i - \bar Y)$, and did regression through the origin.
* If you normalized the data, $\{ \frac{X_i - \bar X}{Sd(X)}, \frac{Y_i - \bar Y}{Sd(Y)}\}$, the slope is $Cor(Y, X)$.


---
## Revisiting Galton's data
### Double check our calculations using R
```{r, fig.height=4,fig.width=4,echo=TRUE}
y <- galton$child
x <- galton$parent
beta1 <- cor(y, x) *  sd(y) / sd(x)
beta0 <- mean(y) - beta1 * mean(x)
rbind(c(beta0, beta1), coef(lm(y ~ x)))
```

---
## Revisiting Galton's data
### Reversing the outcome/predictor relationship
```{r, fig.height=4,fig.width=4,echo=TRUE}
beta1 <- cor(y, x) *  sd(x) / sd(y)
beta0 <- mean(x) - beta1 * mean(y)
rbind(c(beta0, beta1), coef(lm(x ~ y)))
```

---
## Revisiting Galton's data
### Regression through the origin yields an equivalent slope if you center the data first
```{r, fig.height=4,fig.width=4,echo=TRUE}
yc <- y - mean(y)
xc <- x - mean(x)
beta1 <- sum(yc * xc) / sum(xc ^ 2)
c(beta1, coef(lm(y ~ x))[2])
```

---
## Revisiting Galton's data
### Normalizing variables results in the slope being the correlation
```{r, echo=TRUE}
yn <- (y - mean(y))/sd(y)
xn <- (x - mean(x))/sd(x)
c(cor(y, x), cor(yn, xn), coef(lm(yn ~ xn))[2])
```

---
```{r, fig.height=6,fig.width=6,echo=FALSE}
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+20, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq))
g <- g + scale_colour_gradient(low = "lightblue", high="white")  
g <- g + geom_smooth(method="lm", formula=y~x)
g
```


