---
title       : Combining predictors
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, cache=TRUE, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* You can combine classifiers by averaging/voting
* Combining classifiers improves accuracy
* Combining classifiers reduces interpretability
* Boosting, bagging, and random forests are variants on this theme

---

## Netflix prize

BellKor = Combination of 107 predictors 

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/netflix.png height=450>

[http://www.netflixprize.com//leaderboard](http://www.netflixprize.com//leaderboard)

---

## Heritage health prize - Progress Prize 1

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/makers.png height=200>
[Market Makers](https://kaggle2.blob.core.windows.net/wiki-files/327/e4cd1d25-eca9-49ca-9593-b254a773fe03/Market%20Makers%20-%20Milestone%201%20Description%20V2%201.pdf)

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/mestrom.png height=200>

[Mestrom](https://kaggle2.blob.core.windows.net/wiki-files/327/09ccf652-8c1c-4a3d-b979-ce2369c985e4/Willem%20Mestrom%20-%20Milestone%201%20Description%20V2%202.pdf)


---

## Basic intuition - majority vote

Suppose we have 5 completely independent classifiers

If accuracy is 70% for each:
  * $10\times(0.7)^3(0.3)^2 + 5\times(0.7)^4(0.3)^2 + (0.7)^5$
  * 83.7% majority vote accuracy

With 101 independent classifiers
  * 99.9% majority vote accuracy
  

---

## Approaches for combining classifiers

1. Bagging, boosting, random forests
  * Usually combine similar classifiers
2. Combining different classifiers
  * Model stacking
  * Model ensembling

---

## Example with Wage data

__Create training, test and validation sets__

```{r wage}
library(ISLR); data(Wage); library(ggplot2); library(caret);
Wage <- subset(Wage,select=-c(logwage))

# Create a building data set and validation set
inBuild <- createDataPartition(y=Wage$wage,
                              p=0.7, list=FALSE)
validation <- Wage[-inBuild,]; buildData <- Wage[inBuild,]

inTrain <- createDataPartition(y=buildData$wage,
                              p=0.7, list=FALSE)
training <- buildData[inTrain,]; testing <- buildData[-inTrain,]
```


---

## Wage data sets

__Create training, test and validation sets__

```{r, dependson="wage"}
dim(training)
dim(testing)
dim(validation)
```


---

## Build two different models

```{r modFit,dependson="wage"}
mod1 <- train(wage ~.,method="glm",data=training)
mod2 <- train(wage ~.,method="rf",
              data=training, 
              trControl = trainControl(method="cv"),number=3)
```


---

## Predict on the testing set 

```{r predict,dependson="modFit",fig.height=4,fig.width=6}
pred1 <- predict(mod1,testing); pred2 <- predict(mod2,testing)
qplot(pred1,pred2,colour=wage,data=testing)
```


---

## Fit a model that combines predictors

```{r combine,dependson="predict"}
predDF <- data.frame(pred1,pred2,wage=testing$wage)
combModFit <- train(wage ~.,method="gam",data=predDF)
combPred <- predict(combModFit,predDF)
```


---

## Testing errors

```{r ,dependson="combine"}
sqrt(sum((pred1-testing$wage)^2))
sqrt(sum((pred2-testing$wage)^2))
sqrt(sum((combPred-testing$wage)^2))
```


---

## Predict on validation data set

```{r validation,dependson="combine"}
pred1V <- predict(mod1,validation); pred2V <- predict(mod2,validation)
predVDF <- data.frame(pred1=pred1V,pred2=pred2V)
combPredV <- predict(combModFit,predVDF)
```


---

## Evaluate on validation

```{r ,dependson="validation"}
sqrt(sum((pred1V-validation$wage)^2))
sqrt(sum((pred2V-validation$wage)^2))
sqrt(sum((combPredV-validation$wage)^2))
```



---

## Notes and further resources

* Even simple blending can be useful
* Typical model for binary/multiclass data
  * Build an odd number of models
  * Predict with each model
  * Predict the class by majority vote
* This can get dramatically more complicated
  * Simple blending in caret: [caretEnsemble](https://github.com/zachmayer/caretEnsemble) (use at your own risk!)
  * Wikipedia [ensemble learning](http://en.wikipedia.org/wiki/Ensemble_learning)

---

## Recall - scalability matters

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/netflixno.png height=250>
</br></br></br>

[http://www.techdirt.com/blog/innovation/articles/20120409/03412518422/](http://www.techdirt.com/blog/innovation/articles/20120409/03412518422/)

[http://techblog.netflix.com/2012/04/netflix-recommendations-beyond-5-stars.html](http://techblog.netflix.com/2012/04/netflix-recommendations-beyond-5-stars.html)
