---
title       : Exploratory Graphs
subtitle    : 
author      : Roger D. Peng, Associate Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache = T, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Why do we use graphs in data analysis? 

* To understand data properties
* To find patterns in data
* To suggest modeling strategies
* To "debug" analyses
* To communicate results

---

## Exploratory graphs

* <redtext>To understand data properties</redtext>
* <redtext>To find patterns in data</redtext>
* <redtext>To suggest modeling strategies</redtext>
* <redtext>To "debug" analyses</redtext>
* To communicate results

---

## Characteristics of exploratory graphs

* They are made quickly
* A large number are made
* The goal is for personal understanding
* Axes/legends are generally cleaned up (later)
* Color/size are primarily used for information

---


## ACS data


<img class=center src=../../assets/img/acs.png height=350>

[Census ACS Data](http://www.census.gov/acs/www/data_documentation/pums_documentation/)

```{r downloadData, echo=FALSE, cache=TRUE,eval=FALSE}
if(!file.exists("data/ss06pid.csv")) {
        download.file("https://dl.dropboxusercontent.com/u/7710864/data/csv_hid/ss06pid.csv",destfile="data/ss06pid.csv",method="curl")
}
```


```{r loadData,dependson="downloadData"}
pData <- read.csv(gzfile("data/ss06pid.csv.gz"))
```

---

## Before We Start: Summarizing Data


One dimension

* Boxplots
* Histograms
* Barplot
* Density plot
* Five-number summary



---

## Before We Start: Summarizing Data

Two dimensions

* Multiple/overlayed 1-D plots (Lattice/ggplot2)
* Scatterplots
* Smooth scatterplots

$> 2$ dimensions

* Overlayed/multiple 2-D plots; coplots
* Use color, size, shape to add dimensions
* Spinning plots
* Actual 3-D plots (not that useful)


---

## Boxplots

Important parameters: _col_, _varwidth_, _names_, _horizontal_

```{r , dependson="loadData",fig.height=4,fig.width=4}
boxplot(pData$AGEP, col="blue")
```

--- 

## Boxplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
pData <- transform(pData, DDRS = factor(DDRS))
boxplot(AGEP ~ DDRS, data = pData, col = "blue")
```


--- 

## Boxplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
boxplot(AGEP ~ DDRS, data = pData, col = c("blue", "orange"), 
        names = c("yes", "no"), varwidth = TRUE)
```



--- 

## Barplots

```{r , dependson="loadData",fig.height=5,fig.width=5}
barplot(table(pData$CIT), col = "blue")
```



--- 

## Histograms

Important parameters: _breaks_,_freq_,_col_,_xlab_,_ylab_, _xlim, _ylim_ ,_main_

```{r , dependson="loadData",fig.height=4,fig.width=4}
hist(pData$AGEP, col = "blue")
```

--- 

## Histograms

```{r , dependson="loadData",fig.height=5,fig.width=5}
hist(pData$AGEP, col = "blue", breaks = 100, main = "Age")
```


--- 

## Density plots

Important parameters (to plot): _col_,_lwd_,_xlab_,_ylab_,_xlim_,_ylim_

```{r , dependson="loadData",fig.height=4,fig.width=4}
dens <- density(pData$AGEP)
plot(dens, lwd = 3, col = "blue")
```


--- 

## Density plots - multiple distributions

```{r , dependson="loadData",fig.height=4,fig.width=4}
dens <- density(pData$AGEP)
densMales <- density(pData$AGEP[which(pData$SEX == 1)])
plot(dens, lwd = 3, col = "blue")
lines(densMales, lwd = 3, col = "orange")
```

---

## Scatterplots

* Important paramters: _x_, _y_, _type_, _xlab_, _ylab_, _xlim_, _ylim_, _cex_, _col_, _bg_

```{r , dependson="loadData",fig.height=4,fig.width=4,tidy=TRUE}
plot(pData$JWMNP,pData$WAGP,pch=19,col="blue")
```


--- 

## Scatterplots - plotting symbol size


```{r , dependson="loadData",fig.height=5,fig.width=5,tidy=TRUE}
plot(pData$JWMNP,pData$WAGP,pch=19,col="blue",cex=0.5)
```


--- 

## Scatterplots - using color

```{r , dependson="loadData",fig.height=5,fig.width=5,tidy=TRUE}
plot(pData$JWMNP,pData$WAGP,pch=19,col=pData$SEX,cex=0.5)
```

--- 

## Scatterplots - using size

```{r , dependson="loadData",fig.height=5,fig.width=6,tidy=TRUE}
percentMaxAge <- pData$AGEP/max(pData$AGEP)
plot(pData$JWMNP,pData$WAGP,pch=19,col="blue",cex=percentMaxAge*0.5)
```


--- 

## Scatterplots - overlaying lines/points

```{r , dependson="loadData",fig.height=4,fig.width=4, tidy=TRUE}
plot(pData$JWMNP,pData$WAGP,pch=19,col="blue",cex=0.5)
lines(rep(100,dim(pData)[1]),pData$WAGP,col="grey",lwd=5)
points(seq(0,200,length=100),seq(0,20e5,length=100),col="red",pch=19)
```



--- 

## Scatterplots - numeric variables as factors

```{r , dependson="loadData",fig.height=4,fig.width=5}
library(Hmisc)
ageGroups <- cut2(pData$AGEP,g=5)
plot(pData$JWMNP,pData$WAGP,pch=19,col=ageGroups,cex=0.5)
```

---

## If you have a lot of points

```{r,fig.height=4,fig.width=4}
x <- rnorm(1e5)
y <- rnorm(1e5)
plot(x,y,pch=19)
```


---

## If you have a lot of points - sampling

```{r,fig.height=4,fig.width=4}
x <- rnorm(1e5)
y <- rnorm(1e5)
sampledValues <- sample(1:1e5,size=1000,replace=FALSE)
plot(x[sampledValues],y[sampledValues],pch=19)
```


---

## If you have a lot of points - smoothScatter

```{r,fig.height=4,fig.width=4}
x <- rnorm(1e5)
y <- rnorm(1e5)
smoothScatter(x,y)
```



---

## If you have a lot of points - hexbin {hexbin}

```{r,fig.height=4,fig.width=4}
library(hexbin)
x <- rnorm(1e5)
y <- rnorm(1e5)
hbo <- hexbin(x,y)
plot(hbo)
```




---

## QQ-plots

* Important parameters: _x_,_y_

```{r ,fig.height=4,fig.width=4}
x <- rnorm(20); y <- rnorm(20)
qqplot(x,y)
abline(c(0,1))
```


--- 

## Matplot and spaghetti

* Important paramters: _x_, _y_, _lty_,_lwd_,_pch_,_col_

```{r ,fig.height=4,fig.width=4}
X <- matrix(rnorm(20*5),nrow=20)
matplot(X,type="b")
```


--- 

## Heatmaps

* Important paramters: _x_,_y_,_z_,_col_
```{r , dependson="loadData",fig.height=4,fig.width=4}
image(1:10,161:236,as.matrix(pData[1:10,161:236]))
```

---

## Heatmaps - matching intuition

```{r , dependson="loadData", fig.height=4, fig.width=4}
newMatrix <- as.matrix(pData[1:10,161:236])
newMatrix <- t(newMatrix)[,nrow(newMatrix):1]
image(161:236, 1:10, newMatrix)
```


---

## Maps - very basics 

You make need to run `install.packages("maps")` if you don't have the `maps` package installed. 

```{r,dependson="loadData", fig.height=5,fig.width=5}
library(maps)
map("world")
lat <- runif(40,-180,180); lon <- runif(40,-90,90)
points(lat,lon,col="blue",pch=19)
```


---

## Missing values and plots

```{r,fig.height=4,fig.width=4}
x <- c(NA,NA,NA,4,5,6,7,8,9,10)
y <- 1:10
plot(x,y,pch=19,xlim=c(0,11),ylim=c(0,11))
```


---

## Missing values and plots

```{r,fig.height=4,fig.width=4}
x <- rnorm(100)
y <- rnorm(100)
y[x < 0] <- NA
boxplot(x ~ is.na(y))
```


---

## Further resources

* [R Graph Gallery](http://gallery.r-enthusiasts.com/)
* [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html),[ggplot2 basic introduction](http://www.r-bloggers.com/basic-introduction-to-ggplot2/)
* [lattice package](http://cran.r-project.org/web/packages/lattice/index.html),[lattice introduction](http://lmdvr.r-forge.r-project.org/figures/figures.html)
* [R bloggers](http://www.r-bloggers.com/)










