---
title       : Exploratory Analysis Overview
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Exploratory Analysis Content

* Principles of Analytic Graphics
* Exploratory graphs
* Plotting Systems in R
  * base
  * lattice
  * ggplot2
* Hierarchical clustering
* K-Means clustering
* Dimension reduction

---

```{r, echo=FALSE}
library(ggplot2)
```

## Adding a geom

```{r, message=FALSE, fig.height=6, fig.align='center'}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))
```

---

## Principles of Analytic Graphics

* Principle 1: Show comparisons
* Principle 2: Show causality, mechanism, explanation
* Principle 3: Show multivariate data
* Principle 4: Integrate multiple modes of evidence
* Principle 5: Describe and document the evidence
* Principle 6: Content is king

---

## K-means clustering -  example

```{r createData, fig.height=3.5,fig.width=3.5}
set.seed(1234); par(mar=c(0,0,0,0))
x <- rnorm(12,mean=rep(1:3,each=4),sd=0.2)
y <- rnorm(12,mean=rep(c(1,2,1),each=4),sd=0.2)
plot(x,y,col="blue",pch=19,cex=2)
text(x+0.05,y+0.05,labels=as.character(1:12))
```
