l---
title       : Multivariable regression examples
subtitle    : Regression Models
author      : Brian Caffo, Jeff Leek and Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, results='hide'}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
runif(1)
```
## Swiss fertility data
```{r, fig.height=4.5, fig.width=4.5}
library(datasets); data(swiss); require(stats); require(graphics)
pairs(swiss, panel = panel.smooth, main = "Swiss data", col = 3 + (swiss$Catholic > 50))
```

---
## `?swiss`
### Description
Standardized fertility measure and socio-economic indicators for each of 47 French-speaking provinces of Switzerland at about 1888.

A data frame with 47 observations on 6 variables, each of which is in percent, i.e., in [0, 100].

* [,1]   Fertility	Ig, ‘ common standardized fertility measure’
* [,2]	 Agriculture	 % of males involved in agriculture as occupation
* [,3]	 Examination	 % draftees receiving highest mark on army examination
* [,4]	 Education	 % education beyond primary school for draftees.
* [,5]	 Catholic	 % ‘catholic’ (as opposed to ‘protestant’).
* [,6]	 Infant.Mortality	 live births who live less than 1 year.

All variables but ‘Fertility’ give proportions of the population.


---
## Calling `lm`
`summary(lm(Fertility ~ . , data = swiss))`
```{r, echo = FALSE}
summary(lm(Fertility ~ . , data = swiss))$coefficients
```

---
## Example interpretation
* Agriculture is expressed in percentages (0 - 100)
* Estimate is -0.1721.
* We estimate an expected 0.17 decrease in standardized fertility for every 1\% increase in percentage of males involved in agriculture in holding the remaining variables constant.
* The t-test for $H_0: \beta_{Agri} = 0$ versus $H_a: \beta_{Agri} \neq 0$ is  significant.
* Interestingly, the unadjusted estimate is 
```{r}
summary(lm(Fertility ~ Agriculture, data = swiss))$coefficients
```

---
How can adjustment reverse the sign of an effect? Let's try a simulation.
```{r, echo = TRUE}
n <- 100; x2 <- 1 : n; x1 <- .01 * x2 + runif(n, -.1, .1); y = -x1 + x2 + rnorm(n, sd = .01)
summary(lm(y ~ x1))$coef
summary(lm(y ~ x1 + x2))$coef
```

---
```{r, echo = FALSE, fig.height=5, fig.width=10, results = 'show'}
par(mfrow = c(1, 2))
plot(x1, y, pch=21,col="black",bg=topo.colors(n)[x2], frame = FALSE, cex = 1.5)
title('Unadjusted, color is X2')
abline(lm(y ~ x1), lwd = 2)
plot(resid(lm(x1 ~ x2)), resid(lm(y ~ x2)), pch = 21, col = "black", bg = "lightblue", frame = FALSE, cex = 1.5)
title('Adjusted')
abline(0, coef(lm(y ~ x1 + x2))[2], lwd = 2)
```

---
## Back to this data set
* The sign reverses itself with the inclusion of Examination and Education, but of which are negatively correlated with Agriculture.
* The percent of males in the province working in agriculture is negatively related to educational attainment (correlation of `r cor(swiss$Agriculture, swiss$Education)`) and Education and Examination (correlation of `r cor(swiss$Education, swiss$Examination)`) are obviously measuring similar things. 
  * Is the positive marginal an artifact for not having accounted for, say, Education level? (Education does have a stronger effect, by the way.)
* At the minimum, anyone claiming that provinces that are more agricultural have higher fertility rates would immediately be open to criticism.

---
## What if we include an unnecessary variable?
z adds no new linear information, since it's a linear
combination of variables already included. R just drops 
terms that are linear combinations of other terms.
```{r, echo = TRUE}
z <- swiss$Agriculture + swiss$Education
lm(Fertility ~ . + z, data = swiss)
```


---
## Dummy variables are smart
* Consider the linear model
$$
Y_i = \beta_0 + X_{i1} \beta_1 + \epsilon_{i}
$$
where each $X_{i1}$ is binary so that it is a 1 if measurement $i$ is in a group and 0 otherwise. (Treated versus not in a clinical trial, for example.)
* Then for people in the group $E[Y_i] = \beta_0 + \beta_1$
* And for people not in the group $E[Y_i] = \beta_0$
* The LS fits work out to be $\hat \beta_0 + \hat \beta_1$ is the mean for those in the group and $\hat \beta_0$ is the mean for those not in the group.
* $\beta_1$ is interpretted as the increase or decrease in the mean comparing those in the group to those not.
* Note including a binary variable that is 1 for those not in the group would be redundant. It would create three parameters to describe two means.

---
## More than 2 levels
* Consider a multilevel factor level. For didactic reasons, let's say a three level factor (example, US political party affiliation: Republican, Democrat, Independent)
* $Y_i = \beta_0 + X_{i1} \beta_1 + X_{i2} \beta_2 + \epsilon_i$.
* $X_{i1}$ is 1 for Republicans and 0 otherwise.
* $X_{i2}$ is 1 for Democrats and 0 otherwise.
* If $i$ is Republican $E[Y_i] = \beta_0 +\beta_1$
* If $i$ is Democrat $E[Y_i] = \beta_0 + \beta_2$.
* If $i$ is Independent $E[Y_i] = \beta_0$. 
* $\beta_1$ compares Republicans to Independents.
* $\beta_2$ compares Democrats to Independents.
* $\beta_1 - \beta_2$ compares Republicans to Democrats.
* (Choice of reference category changes the interpretation.)

---
## Insect Sprays
```{r, echo = FALSE, fig.height=5, fig.width=5}
require(datasets);data(InsectSprays)
require(stats); require(graphics)
boxplot(count ~ spray, data = InsectSprays,
        xlab = "Type of spray", ylab = "Insect count",
        main = "InsectSprays data", varwidth = TRUE, col = "lightgray")
```

---
## Linear model fit, group A is the reference
```{r, echo= TRUE}
summary(lm(count ~ spray, data = InsectSprays))$coef
```

---
## Hard coding the dummy variables
```{r, echo= TRUE}
summary(lm(count ~ 
             I(1 * (spray == 'B')) + I(1 * (spray == 'C')) + 
             I(1 * (spray == 'D')) + I(1 * (spray == 'E')) +
             I(1 * (spray == 'F'))
           , data = InsectSprays))$coef
```

---
## What if we include all 6?
```{r, echo= TRUE}
lm(count ~ 
   I(1 * (spray == 'B')) + I(1 * (spray == 'C')) +  
   I(1 * (spray == 'D')) + I(1 * (spray == 'E')) +
   I(1 * (spray == 'F')) + I(1 * (spray == 'A')), data = InsectSprays)
```

---
## What if we omit the intercept?
```{r, echo= TRUE}
summary(lm(count ~ spray - 1, data = InsectSprays))$coef
unique(ave(InsectSprays$count, InsectSprays$spray))
```

---
## Summary
* If we treat Spray as a factor, R includes an intercept and omits the alphabetically first level of the factor.
  * All t-tests are for comparisons of Sprays versus Spray A.
  * Emprirical mean for A is the intercept.
  * Other group means are the itc plus their coefficient. 
* If we omit an intercept, then it includes terms for all levels of the factor. 
  * Group means are the coefficients. 
  * Tests are tests of whether the groups are different than zero. (Are the expected counts zero for that spray.)
* If we want comparisons between, Spray B and C, say we could refit the model with C (or B) as the reference level. 

---
## Reordering the levels
```{r}
spray2 <- relevel(InsectSprays$spray, "C")
summary(lm(count ~ spray2, data = InsectSprays))$coef
```

---
## Doing it manually
Equivalently 
$$Var(\hat \beta_B - \hat \beta_C) = Var(\hat \beta_B) + Var(\hat \beta_C) - 2 Cov(\hat \beta_B, \hat \beta_C)$$
```{r}
fit <- lm(count ~ spray, data = InsectSprays) #A is ref
bbmbc <- coef(fit)[2] - coef(fit)[3] #B - C
temp <- summary(fit) 
se <- temp$sigma * sqrt(temp$cov.unscaled[2, 2] + temp$cov.unscaled[3,3] - 2 *temp$cov.unscaled[2,3])
t <- (bbmbc) / se
p <- pt(-abs(t), df = fit$df)
out <- c(bbmbc, se, t, p)
names(out) <- c("B - C", "SE", "T", "P")
round(out, 3)
```

---
## Other thoughts on this data
* Counts are bounded from below by 0, violates the assumption of normality of the errors. 
  * Also there are counts near zero, so both the actual assumption and the intent of the assumption are violated.
* Variance does not appear to be constant.
* Perhaps taking logs of the counts would help. 
  * There are 0 counts, so maybe log(Count + 1)
* Also, we'll cover Poisson GLMs for fitting count data.

---

## Example - Millenium Development Goal 1

[http://www.un.org/millenniumgoals/pdf/MDG_FS_1_EN.pdf](http://www.un.org/millenniumgoals/pdf/MDG_FS_1_EN.pdf)

[http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*](http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*)

---

## WHO childhood hunger data


```{r whoDataLoad}
#download.file("http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*","hunger.csv",method="curl")
hunger <- read.csv("hunger.csv")
hunger <- hunger[hunger$Sex!="Both sexes",]
head(hunger)
```

---

## Plot percent hungry versus time

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
```

---

## Remember the linear model

$$Hu_i = b_0 + b_1 Y_i + e_i$$

$b_0$ = percent hungry at Year 0

$b_1$ = decrease in percent hungry per year

$e_i$ = everything we didn't measure

---

## Add the linear model

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
lines(hunger$Year,lm1$fitted,lwd=3,col="darkgrey")
```


---

## Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
```

---
 
## Now two lines

$$HuF_i = bf_0 + bf_1 YF_i + ef_i$$

$bf_0$ = percent of girls hungry at Year 0

$bf_1$ = decrease in percent of girls hungry per year

$ef_i$ = everything we didn't measure 


$$HuM_i = bm_0 + bm_1 YM_i + em_i$$

$bm_0$ = percent of boys hungry at Year 0

$bm_1$ = decrease in percent of boys hungry per year

$em_i$ = everything we didn't measure 



---

## Color by male/female

```{r, dependson="whoDataLoad",fig.height=3.5,fig.width=4}
lmM <- lm(hunger$Numeric[hunger$Sex=="Male"] ~ hunger$Year[hunger$Sex=="Male"])
lmF <- lm(hunger$Numeric[hunger$Sex=="Female"] ~ hunger$Year[hunger$Sex=="Female"])
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
lines(hunger$Year[hunger$Sex=="Male"],lmM$fitted,col="black",lwd=3)
lines(hunger$Year[hunger$Sex=="Female"],lmF$fitted,col="red",lwd=3)
```


---

## Two lines, same slope

$$Hu_i = b_0 + b_1 \mathbb{1}(Sex_i="Male") + b_2 Y_i + e^*_i$$

$b_0$ - percent hungry at year zero for females

$b_0 + b_1$ - percent hungry at year zero for males

$b_2$ - change in percent hungry (for either males or females) in one year

$e^*_i$ - everything we didn't measure

---

## Two lines, same slope in R


```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] ),col="black",lwd=3)
```

---

## Two lines, different slopes (interactions)

$$Hu_i = b_0 + b_1 \mathbb{1}(Sex_i="Male") + b_2 Y_i + b_3 \mathbb{1}(Sex_i="Male")\times Y_i + e^+_i$$

$b_0$ - percent hungry at year zero for females

$b_0 + b_1$ - percent hungry at year zero for males

$b_2$ - change in percent hungry (females) in one year

$b_2 + b_3$ - change in percent hungry (males) in one year

$e^+_i$ - everything we didn't measure

---

## Two lines, different slopes in R


```{r lmBothChunk, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex + hunger$Sex*hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] +lmBoth$coeff[4]),col="black",lwd=3)
```


---

## Two lines, different slopes in R


```{r, dependson="lmBothChunk",fig.height=4,fig.width=4}
summary(lmBoth)
```

---
## Interpretting a continuous interaction
$$
E[Y_i | X_{1i}=x_1, X_{2i}=x_2] = \beta_0 + \beta_1 x_{1} + \beta_2 x_{2} + \beta_3 x_{1}x_{2}
$$
Holding $X_2$ constant we have
$$
E[Y_i | X_{1i}=x_1+1, X_{2i}=x_2]-E[Y_i | X_{1i}=x_1, X_{2i}=x_2]
= \beta_1 + \beta_3 x_{2} 
$$
And thus the expected change in $Y$ per unit change in $X_1$ holding all else constant is not constant. $\beta_1$ is the slope when $x_{2} = 0$. Note further that:
$$
E[Y_i | X_{1i}=x_1+1, X_{2i}=x_2+1]-E[Y_i | X_{1i}=x_1, X_{2i}=x_2+1]
$$
$$
-E[Y_i | X_{1i}=x_1+1, X_{2i}=x_2]-E[Y_i | X_{1i}=x_1, X_{2i}=x_2]
$$
$$
=\beta_3  
$$
Thus, $\beta_3$ is the change in the expected change in $Y$ per unit change in $X_1$, per unit change in $X_2$.

Or, the change in the slope relating $X_1$ and $Y$ per unit change in $X_2$.

---

## Example

$$Hu_i = b_0 + b_1 In_i + b_2 Y_i + b_3 In_i \times Y_i + e^+_i$$

$b_0$ - percent hungry at year zero for children with whose parents have no income

$b_1$ - change in percent hungry for each dollar of income in year zero

$b_2$ - change in percent hungry in one year for children whose parents have no income

$b_3$ - increased change in percent hungry by year for each dollar of income  - e.g. if income is $10,000, then change in percent hungry in one year will be

$$b_2 + 1e4 \times b_3$$

$e^+_i$ - everything we didn't measure

__Lot's of care/caution needed!__


