---
title       : rCharts
subtitle    : Building Data Products
author      : Brian Caffo, Jeff Leek, Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012
highlighter : highlight.js  
hitheme     : tomorrow       
url:
    lib: ../../librariesNew #Remove new if using old slidify
    assets: ../../assets
widgets     : [mathjax, quiz, bootstrap]
mode        : selfcontained # {standalone, draft}
ext_widgets : {rCharts: ["libraries/nvd3"]}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, results='hide'}
# make this an external chunk that can be included in any file
library(knitr)
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
runif(1)
```

## rCharts
- rCharts is a way to create interactive javascript visualizations using R
- So
  - You don't have to learn complex tools, like D3
  - You simply work in R learning a minimal amount of new syntaxt
- rCharts was written by Ramnath Vaidyanathan (friend of the Data Science Series), who also wrote slidify, the framework we use for all of the lectures in the class
- This lecture is basically going through
  (http://ramnathv.github.io/rCharts/)

---
## Example
```
require(rCharts)
haireye = as.data.frame(HairEyeColor)
nPlot(Freq ~ Hair, group = 'Eye', type = 'multiBarChart',
  data = subset(haireye, Sex == 'Male')
)
```

---
## Holy eye candy Batman!
```{r, echo = FALSE, results = 'hide'}
require(rCharts)
```
```{r echo = FALSE, results = 'asis'}
haireye = as.data.frame(HairEyeColor)
n1 <- nPlot(Freq ~ Hair, group = 'Eye', type = 'multiBarChart',
  data = subset(haireye, Sex == 'Male')
)
n1$print('chart1')
```

---
## Slidify interactive
- The above was an example of embedding an rChart in a slidify document
  - In the YAML 
```yaml ext_widgets : {rCharts: ["libraries/nvd3"]}```

---
## Deconstructing another example
```
## Example 1 Facetted Scatterplot
names(iris) = gsub("\\.", "", names(iris))
rPlot(SepalLength ~ SepalWidth | Species, data = iris, color = 'Species', type = 'point')
```


---
## When run
```{r echo = FALSE, results = 'asis'}
## Example 1 Facetted Scatterplot
names(iris) = gsub("\\.", "", names(iris))
rPlot(SepalLength ~ SepalWidth | Species, data = iris, color = 'Species', type = 'point')
```

---
## Example 2 Facetted Barplot
```
hair_eye = as.data.frame(HairEyeColor)
rPlot(Freq ~ Hair | Eye, color = 'Eye', data = hair_eye, type = 'bar')
```

---
## Example 2 Facetted Barplot, when run
```{r echo = FALSE, results = 'asis'}
hair_eye = as.data.frame(HairEyeColor)
rPlot(Freq ~ Hair | Eye, color = 'Eye', data = hair_eye, type = 'bar')
```


