---
title       : Downloading files
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---

## Get/set your working directory

* A basic component of working with data is knowing your working directory
* The two main commands are ```getwd()``` and ```setwd()```. 
* Be aware of relative versus absolute paths
  * __Relative__ - ```setwd("./data")```, ```setwd("../")```
  * __Absolute__ - ```setwd("/Users/jtleek/data/")```
* Important difference in Windows ```setwd("C:\\Users\\Andrew\\Downloads")```

---

## Checking for and creating directories

* ```file.exists("directoryName")``` will check to see if the directory exists
* ```dir.create("directoryName")``` will create a directory if it doesn't exist
* Here is an example checking for a "data" directory and creating it if it doesn't exist

```{r data}
if(!file.exists("data")){
  dir.create("data")
}
```


---

## Getting data from the internet - download.file()

* Downloads a file from the internet
* Even if you could do this by hand, helps with reproducibility
* Important parameters are _url_, _destfile_, _method_
* Useful for downloading tab-delimited, csv, and other files


---

## Example - Baltimore camera data


<img class=center src=../../assets/img/03_ObtainingData/cameras.png height=500>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)


---

## Example - Baltimore camera data

<img class=center src=../../assets/img/03_ObtainingData/cameraslink.png height=500>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)


---

## Download a file from the web

```{r,dependson="data"}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.csv",method="curl")
list.files("./data")
dateDownloaded <- date()
dateDownloaded
```

---

## Some notes about download.file()

* If the url starts with _http_ you can use download.file()
* If the url starts with _https_ on Windows you may be ok
* If the url starts with _https_ on Mac you may need to set _method="curl"_
* If the file is big, this might take a while
* Be sure to record when you downloaded. 
