---
title       : Some Common Distributions
subtitle    : Statistical Inference
author      : Brian Caffo, Jeff Leek, Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


## The Bernoulli distribution

- The **Bernoulli distribution** arises as the result of a binary outcome
- Bernoulli random variables take (only) the values 1 and 0 with probabilities of (say) $p$ and $1-p$ respectively
- The PMF for a Bernoulli random variable $X$ is $$P(X = x) =  p^x (1 - p)^{1 - x}$$
- The mean of a Bernoulli random variable is $p$ and the variance is $p(1 - p)$
- If we let $X$ be a Bernoulli random variable, it is typical to call $X=1$ as a "success" and $X=0$ as a "failure"


---

## Binomial trials

- The *binomial random variables* are obtained as the sum of iid Bernoulli trials
- In specific, let $X_1,\ldots,X_n$ be iid Bernoulli$(p)$; then $X = \sum_{i=1}^n X_i$ is a binomial random variable
- The binomial mass function is
$$
P(X = x) = 
\left(
\begin{array}{c}
  n \\ x
\end{array}
\right)
p^x(1 - p)^{n-x}
$$
for $x=0,\ldots,n$

---

## Choose

- Recall that the notation 
  $$\left(
    \begin{array}{c}
      n \\ x
    \end{array}
  \right) = \frac{n!}{x!(n-x)!}
  $$ (read "$n$ choose $x$") counts the number of ways of selecting $x$ items out of $n$
  without replacement disregarding the order of the items

$$\left(
    \begin{array}{c}
      n \\ 0
    \end{array}
  \right) =
\left(
    \begin{array}{c}
      n \\ n
    \end{array}
  \right) =  1
  $$ 

---

## Example

- Suppose a friend has $8$ children (oh my!), $7$ of which are girls and none are twins
- If each gender has an independent $50$% probability for each birth, what's the probability of getting $7$ or more girls out of $8$ births?
$$\left(
\begin{array}{c}
  8 \\ 7
\end{array}
\right) .5^{7}(1-.5)^{1}
+
\left(
\begin{array}{c}
  8 \\ 8
\end{array}
\right) .5^{8}(1-.5)^{0} \approx 0.04
$$
```{r}
choose(8, 7) * .5 ^ 8 + choose(8, 8) * .5 ^ 8 
pbinom(6, size = 8, prob = .5, lower.tail = FALSE)
```


---

## The normal distribution

- A random variable is said to follow a **normal** or **Gaussian** distribution with mean $\mu$ and variance $\sigma^2$ if the associated density is
  $$
  (2\pi \sigma^2)^{-1/2}e^{-(x - \mu)^2/2\sigma^2}
  $$
  If $X$ a RV with this density then $E[X] = \mu$ and $Var(X) = \sigma^2$
- We write $X\sim \mbox{N}(\mu, \sigma^2)$
- When $\mu = 0$ and $\sigma = 1$ the resulting distribution is called **the standard normal distribution**
- Standard normal RVs are often labeled $Z$

---
## The standard normal distribution with reference lines 
```{r, fig.height=6, fig.width=6, fig.align='center', echo = FALSE}
x <- seq(-3, 3, length = 1000)
library(ggplot2)
g <- ggplot(data.frame(x = x, y = dnorm(x)), 
            aes(x = x, y = y)) + geom_line(size = 2)
g <- g + geom_vline(xintercept = -3 : 3, size = 2)
g
```

---

## Facts about the normal density

If $X \sim \mbox{N}(\mu,\sigma^2)$ then 
$$Z = \frac{X -\mu}{\sigma} \sim N(0, 1)$$ 


If $Z$ is standard normal $$X = \mu + \sigma Z \sim \mbox{N}(\mu, \sigma^2)$$

---

## More facts about the normal density

1. Approximately $68\%$, $95\%$ and $99\%$  of the normal density lies within $1$, $2$ and $3$ standard deviations from the mean, respectively
2. $-1.28$, $-1.645$, $-1.96$ and $-2.33$ are the $10^{th}$, $5^{th}$, $2.5^{th}$ and $1^{st}$ percentiles of the standard normal distribution respectively
3. By symmetry, $1.28$, $1.645$, $1.96$ and $2.33$ are the $90^{th}$, $95^{th}$, $97.5^{th}$ and $99^{th}$ percentiles of the standard normal distribution respectively

---

## Question

- What is the $95^{th}$ percentile of a $N(\mu, \sigma^2)$ distribution? 
  - Quick answer in R `qnorm(.95, mean = mu, sd = sd)`
- Or, because you have the standard normal quantiles memorized
and you know that 1.645 is the 95th percentile you know that the answer has to be
$$\mu + \sigma 1.645$$
- (In general $\mu + \sigma z_0$ where $z_0$ is the appropriate standard normal quantile)

---

## Question

- What is the probability that a $\mbox{N}(\mu,\sigma^2)$ RV is larger than $x$?

---
## Example

Assume that the number of daily ad clicks for a company 
is (approximately) normally distributed with a mean of 1020 and a standard
deviation of 50. What's the probability of getting
more than  1,160 clicks in a day?

---

## Example

Assume that the number of daily ad clicks for a company 
is (approximately) normally distributed with a mean of 1020 and a standard
deviation of 50. What's the probability of getting
more than  1,160 clicks in a day?

It's not very likely, 1,160 is `r (1160 - 1020) / 50` standard
deviations from the mean 
```{r}
pnorm(1160, mean = 1020, sd = 50, lower.tail = FALSE)
pnorm(2.8, lower.tail = FALSE)
```

---

## Example

Assume that the number of daily ad clicks for a company 
is (approximately) normally distributed with a mean of 1020 and a standard
deviation of 50. What number of daily ad clicks would represent
the one where 75% of days have fewer clicks (assuming
days are independent and identically distributed)?

---

## Example

Assume that the number of daily ad clicks for a company 
is (approximately) normally distributed with a mean of 1020 and a standard
deviation of 50. What number of daily ad clicks would represent
the one where 75% of days have fewer clicks (assuming
days are independent and identically distributed)?

```{r}
qnorm(0.75, mean = 1020, sd = 50)
```

---
## The Poisson distribution
* Used to model counts
* The Poisson mass function is
$$
P(X = x; \lambda) = \frac{\lambda^x e^{-\lambda}}{x!}
$$
for $x=0,1,\ldots$
* The mean of this distribution is $\lambda$
* The variance of this distribution is $\lambda$
* Notice that $x$ ranges from $0$ to $\infty$

---
## Some uses for the Poisson distribution
* Modeling count data  
* Modeling event-time or survival data
* Modeling contingency tables
* Approximating binomials when $n$ is large and $p$ is small

---
## Rates and Poisson random variables
* Poisson random variables are used to model rates
* $X \sim Poisson(\lambda t)$ where 
  * $\lambda = E[X / t]$ is the expected count per unit of time
  * $t$ is the total monitoring time

---
## Example
The number of people that show up at a bus stop is Poisson with
a mean of $2.5$ per hour.

If watching the bus stop for 4 hours, what is the probability that $3$
or fewer people show up for the whole time?

```{r}
ppois(3, lambda = 2.5 * 4)
```

---
## Poisson approximation to the binomial
* When $n$ is large and $p$ is small the Poisson distribution
  is an accurate approximation to the binomial distribution
* Notation
  * $X \sim \mbox{Binomial}(n, p)$
  * $\lambda = n p$
  * $n$ gets large 
  * $p$ gets small


---
## Example, Poisson approximation to the binomial

We flip a coin with success probablity $0.01$ five hundred times. 

What's the probability of 2 or fewer successes?

```{r}
pbinom(2, size = 500, prob = .01)
ppois(2, lambda=500 * .01)
```

