---
title       : Multiple regression
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Key ideas

* Regression with multiple covariates
* Still using least squares/central limit theorem
* Interpretation depends on all variables


---

## Example - Millenium Development Goal 1

<img class=center src=../../assets/img/mdg1.png height=450>

[http://www.un.org/millenniumgoals/pdf/MDG_FS_1_EN.pdf](http://www.un.org/millenniumgoals/pdf/MDG_FS_1_EN.pdf)

[http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*](http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*)

---

## WHO childhood hunger data


```{r whoDataLoad}
download.file("http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*","./data/hunger.csv",method="curl")
hunger <- read.csv("./data/hunger.csv")
hunger <- hunger[hunger$Sex!="Both sexes",]
head(hunger)
```


---

## Plot percent hungry versus time

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
```


---

## Remember the linear model

$$Hu_i = b_0 + b_1 Y_i + e_i$$

$b_0$ = percent hungry at Year 0

$b_1$ = decrease in percent hungry per year

$e_i$ = everything we didn't measure

---

## Add the linear model

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
lines(hunger$Year,lm1$fitted,lwd=3,col="darkgrey")
```


---

## Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
```



---

## Now two lines

$$HuF_i = bf_0 + bf_1 YF_i + ef_i$$

$bf_0$ = percent of girls hungry at Year 0

$bf_1$ = decrease in percent of girls hungry per year

$ef_i$ = everything we didn't measure 


$$HuM_i = bm_0 + bm_1 YM_i + em_i$$

$bm_0$ = percent of boys hungry at Year 0

$bm_1$ = decrease in percent of boys hungry per year

$em_i$ = everything we didn't measure 



---

## Color by male/female

```{r, dependson="whoDataLoad",fig.height=3.5,fig.width=4}
lmM <- lm(hunger$Numeric[hunger$Sex=="Male"] ~ hunger$Year[hunger$Sex=="Male"])
lmF <- lm(hunger$Numeric[hunger$Sex=="Female"] ~ hunger$Year[hunger$Sex=="Female"])
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
lines(hunger$Year[hunger$Sex=="Male"],lmM$fitted,col="black",lwd=3)
lines(hunger$Year[hunger$Sex=="Female"],lmF$fitted,col="red",lwd=3)
```


---

## Two lines, same slope

$$Hu_i = b_0 + b_1 \mathbb{1}(Sex_i="Male") + b_2 Y_i + e^*_i$$

$b_0$ - percent hungry at year zero for females

$b_0 + b_1$ - percent hungry at year zero for males

$b_2$ - change in percent hungry (for either males or females) in one year

$e^*_i$ - everything we didn't measure

---

## Two lines, same slope in R


```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] ),col="black",lwd=3)
```





---

## Two lines, different slopes (interactions)

$$Hu_i = b_0 + b_1 \mathbb{1}(Sex_i="Male") + b_2 Y_i + b_3 \mathbb{1}(Sex_i="Male")\times Y_i + e^+_i$$

$b_0$ - percent hungry at year zero for females

$b_0 + b_1$ - percent hungry at year zero for males

$b_2$ - change in percent hungry (females) in one year

$b_2 + b_3$ - change in percent hungry (males) in one year

$e^+_i$ - everything we didn't measure

---

## Two lines, different slopes in R


```{r lmBothChunk, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex + hunger$Sex*hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] +lmBoth$coeff[4]),col="black",lwd=3)
```


---

## Two lines, different slopes in R


```{r, dependson="lmBothChunk",fig.height=4,fig.width=4}
summary(lmBoth)
```

---

## Interactions for continuous variables

$$Hu_i = b_0 + b_1 In_i + b_2 Y_i + b_3 In_i \times Y_i + e^+_i$$

$b_0$ - percent hungry at year zero for children with whose parents have no income

$b_1$ - change in percent hungry for each dollar of income in year zero

$b_2$ - change in percent hungry in one year for children whose parents have no income

$b_3$ - increased change in percent hungry by year for each dollar of income  - e.g. if income is $10,000, then change in percent hungry in one year will be

$$b_2 + 1e4 \times b_3$$

$e^+_i$ - everything we didn't measure

__Lot's of care/caution needed!__



