---
title       : Reading Excel files
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=TRUE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```



## Excel files

_Still probably the most widely used format for sharing data_

<img class=center src=../../assets/img/03_ObtainingData/excel2.png height=450>


[http://office.microsoft.com/en-us/excel/](http://office.microsoft.com/en-us/excel/)

---


## Example - Baltimore camera data

<img class=center src=../../assets/img/03_ObtainingData/cameras.png height=500>

[https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru](https://data.baltimorecity.gov/Transportation/Baltimore-Fixed-Speed-Cameras/dz54-2aru)

---

## Download the file to load

```{r data}
if(!file.exists("data")){dir.create("data")}
fileUrl <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.xlsx?accessType=DOWNLOAD"
download.file(fileUrl,destfile="./data/cameras.xlsx",method="curl")
dateDownloaded <- date()
```

---

## read.xlsx(), read.xlsx2() {xlsx package}

```{r xlsx,dependson="data"}
library(xlsx)
cameraData <- read.xlsx("./data/cameras.xlsx",sheetIndex=1,header=TRUE)
head(cameraData)
```


---

## Reading specific rows and columns

```{r ,dependson="xlsx"}
colIndex <- 2:3
rowIndex <- 1:4
cameraDataSubset <- read.xlsx("./data/cameras.xlsx",sheetIndex=1,
                              colIndex=colIndex,rowIndex=rowIndex)
cameraDataSubset
```

---

## Further notes

* The _write.xlsx_ function will write out an Excel file with similar arguments.
* _read.xlsx2_ is much faster than _read.xlsx_ but for reading subsets of rows may be slightly unstable. 
* The [XLConnect](http://cran.r-project.org/web/packages/XLConnect/index.html) package has more options for writing and manipulating Excel files
* The [XLConnect vignette](http://cran.r-project.org/web/packages/XLConnect/vignettes/XLConnect.pdf) is a good place to start for that package
* In general it is advised to store your data in either a database
or in comma separated files (.csv) or tab separated files (.tab/.txt) as they are easier to distribute.
