---
title       : The caret package
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## The caret R package

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/caret.png height=400>

[http://caret.r-forge.r-project.org/](http://caret.r-forge.r-project.org/)

---

## Caret functionality

* Some preprocessing (cleaning)
  * preProcess
* Data splitting
  * createDataPartition
  * createResample
  * createTimeSlices
* Training/testing functions
  * train
  * predict
* Model comparison
  * confusionMatrix

---

## Machine learning algorithms in R

* Linear discriminant analysis
* Regression
* Naive Bayes
* Support vector machines
* Classification and regression trees
* Random forests
* Boosting
* etc. 

---

## Why caret? 

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/predicttable.png height=250>

http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf


--- 

## SPAM Example: Data splitting

```{r loadPackage}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```


--- 

## SPAM Example: Fit a model

```{r training, dependson="loadPackage",cache=TRUE}
set.seed(32343)
modelFit <- train(type ~.,data=training, method="glm")
modelFit
```


--- 

## SPAM Example: Final model

```{r finalModel, dependson="training",cache=TRUE}
modelFit <- train(type ~.,data=training, method="glm")
modelFit$finalModel
```


--- 

## SPAM Example: Prediction

```{r predictions, dependson="training",cache=TRUE}
predictions <- predict(modelFit,newdata=testing)
predictions
```

--- 

## SPAM Example: Confusion Matrix

```{r confusion, dependson="predictions",cache=TRUE}
confusionMatrix(predictions,testing$type)
```

---

## Further information

* Caret tutorials:
  * [http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf](http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf)
  * [http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf](http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf)
* A paper introducing the caret package
  * [http://www.jstatsoft.org/v28/i05/paper](http://www.jstatsoft.org/v28/i05/paper)