---
title       : Reading HDF5
subtitle    : 
author      : Jeffrey Leek 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/', cache=FALSE)

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## HDF5

* Used for storing large data sets
* Supports storing a range of data types
* Hierarchical data format
* _groups_ containing zero or more data sets and metadata
  * Have a _group header_ with group name and list of attributes
  * Have a _group symbol table_ with a list of objects in group
* _datasets_ multidimensional array of data elements with metadata
  * Have a _header_ with name, datatype, dataspace, and storage layout
  * Have a _data array_ with the data

[http://www.hdfgroup.org/](http://www.hdfgroup.org/)

---


## R HDF5 package

```{r removeExample,echo=FALSE,message=FALSE}
invisible(if(file.exists("example.h5")){file.remove("example.h5")})
```

```{r loadPackage,eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("rhdf5")
```

```{r rhdf5,message=FALSE}
library(rhdf5)
created = h5createFile("example.h5")
created
```

* This will install packages from Bioconductor [http://bioconductor.org/](http://bioconductor.org/), primarily used for genomics but also has good "big data" packages
* Can be used to interface with hdf5 data sets. 
* This lecture is modeled very closely on the rhdf5 tutorial that
can be found here [http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf](http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf)

--- 

## Create groups

```{r groups,dependson="rhdf5"}
created = h5createGroup("example.h5","foo")
created = h5createGroup("example.h5","baa")
created = h5createGroup("example.h5","foo/foobaa")
h5ls("example.h5")
```

---

## Write to groups

```{r wgroups,dependson="groups"}
A = matrix(1:10,nr=5,nc=2)
h5write(A, "example.h5","foo/A")
B = array(seq(0.1,2.0,by=0.1),dim=c(5,2,2))
attr(B, "scale") <- "liter"
h5write(B, "example.h5","foo/foobaa/B")
h5ls("example.h5")
```

---

## Write a data set 

```{r dataset,dependson="wgroups"}
df = data.frame(1L:5L,seq(0,1,length.out=5),
  c("ab","cde","fghi","a","s"), stringsAsFactors=FALSE)
h5write(df, "example.h5","df")
h5ls("example.h5")
```

---

## Reading data

```{r read,dependson="dataset"}
readA = h5read("example.h5","foo/A")
readB = h5read("example.h5","foo/foobaa/B")
readdf= h5read("example.h5","df")
readA
```


---

## Writing and reading chunks

```{r ,dependson="read"}
h5write(c(12,13,14),"example.h5","foo/A",index=list(1:3,1))
h5read("example.h5","foo/A")
```

---

## Notes and further resources

* hdf5 can be used to optimize reading/writing from disc in R
* The rhdf5 tutorial: 
  * [http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf](http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf)
* The HDF group has informaton on HDF5 in general [http://www.hdfgroup.org/HDF5/](http://www.hdfgroup.org/HDF5/)
