---
title       : Multiple testing
subtitle    : Statistical Inference 
author      : Brian Caffo, Jeffrey Leek, Roger Peng
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
## Key ideas

* Hypothesis testing/significance analysis is commonly overused
* Correcting for multiple testing avoids false positives or discoveries
* Two key components
  * Error measure
  * Correction


---

## Three eras of statistics

__The age of Quetelet and his successors, in which huge census-level data sets were brought to bear on simple but important questions__: Are there more male than female births? Is the rate of insanity rising?

The classical period of Pearson, Fisher, Neyman, Hotelling, and their successors, intellectual giants who __developed a theory of optimal inference capable of wringing every drop of information out of a scientific experiment__. The questions dealt with still tended to be simple Is treatment A better than treatment B? 

__The era of scientific mass production__, in which new technologies typified by the microarray allow a single team of scientists to produce data sets of a size Quetelet would envy. But now the flood of data is accompanied by a deluge of questions, perhaps thousands of estimates or hypothesis tests that the statistician is charged with answering together; not at all what the classical masters had in mind. Which variables matter among the thousands measured? How do you relate unrelated information?

[http://www-stat.stanford.edu/~ckirby/brad/papers/2010LSIexcerpt.pdf](http://www-stat.stanford.edu/~ckirby/brad/papers/2010LSIexcerpt.pdf)

---

## Reasons for multiple testing

<img class=center src=fig/datasources.png height=450>


---

## Why correct for multiple tests?

<img class=center src=fig/jellybeans1.png height=450>


[http://xkcd.com/882/](http://xkcd.com/882/)

---

## Why correct for multiple tests?

<img class=center src=fig/jellybeans2.png height=400>

[http://xkcd.com/882/](http://xkcd.com/882/)


---

## Types of errors

Suppose you are testing a hypothesis that a parameter $\beta$ equals zero versus the alternative that it does not equal zero. These are the possible outcomes. 
</br></br>

                    | $\beta=0$   | $\beta\neq0$   |  Hypotheses
--------------------|-------------|----------------|---------
Claim $\beta=0$     |      $U$    |      $T$       |  $m-R$
Claim $\beta\neq 0$ |      $V$    |      $S$       |  $R$
    Claims          |     $m_0$   |      $m-m_0$   |  $m$

</br></br>

__Type I error or false positive ($V$)__ Say that the parameter does not equal zero when it does

__Type II error or false negative ($T$)__ Say that the parameter equals zero when it doesn't 


---

## Error rates

__False positive rate__ - The rate at which false results ($\beta = 0$) are called significant: $E\left[\frac{V}{m_0}\right]$*

__Family wise error rate (FWER)__ - The probability of at least one false positive ${\rm Pr}(V \geq 1)$

__False discovery rate (FDR)__ - The rate at which claims of significance are false $E\left[\frac{V}{R}\right]$

* The false positive rate is closely related to the type I error rate [http://en.wikipedia.org/wiki/False_positive_rate](http://en.wikipedia.org/wiki/False_positive_rate)

---

## Controlling the false positive rate

If P-values are correctly calculated calling all $P < \alpha$ significant will control the false positive rate at level $\alpha$ on average. 

<redtext>Problem</redtext>: Suppose that you perform 10,000 tests and $\beta = 0$ for all of them. 

Suppose that you call all $P < 0.05$ significant. 

The expected number of false positives is: $10,000 \times 0.05 = 500$  false positives. 

__How do we avoid so many false positives?__


---

## Controlling family-wise error rate (FWER)


The [Bonferroni correction](http://en.wikipedia.org/wiki/Bonferroni_correction) is the oldest multiple testing correction. 

__Basic idea__: 
* Suppose you do $m$ tests
* You want to control FWER at level $\alpha$ so $Pr(V \geq 1) < \alpha$
* Calculate P-values normally
* Set $\alpha_{fwer} = \alpha/m$
* Call all $P$-values less than $\alpha_{fwer}$ significant

__Pros__: Easy to calculate, conservative
__Cons__: May be very conservative


---

## Controlling false discovery rate (FDR)

This is the most popular correction when performing _lots_ of tests say in genomics, imaging, astronomy, or other signal-processing disciplines. 

__Basic idea__: 
* Suppose you do $m$ tests
* You want to control FDR at level $\alpha$ so $E\left[\frac{V}{R}\right]$
* Calculate P-values normally
* Order the P-values from smallest to largest $P_{(1)},...,P_{(m)}$
* Call any $P_{(i)} \leq \alpha \times \frac{i}{m}$ significant

__Pros__: Still pretty easy to calculate, less conservative (maybe much less)

__Cons__: Allows for more false positives, may behave strangely under dependence

---

## Example with 10 P-values

<img class=center src=fig/example10pvals.png height=450>

Controlling all error rates at $\alpha = 0.20$

---

## Adjusted P-values

* One approach is to adjust the threshold $\alpha$
* A different approach is to calculate "adjusted p-values"
* They _are not p-values_ anymore
* But they can be used directly without adjusting $\alpha$

__Example__: 
* Suppose P-values are $P_1,\ldots,P_m$
* You could adjust them by taking $P_i^{fwer} = \max{m \times P_i,1}$ for each P-value.
* Then if you call all $P_i^{fwer} < \alpha$ significant you will control the FWER. 

---

## Case study I: no true positives

```{r createPvals,cache=TRUE}
set.seed(1010093)
pValues <- rep(NA,1000)
for(i in 1:1000){
  y <- rnorm(20)
  x <- rnorm(20)
  pValues[i] <- summary(lm(y ~ x))$coeff[2,4]
}

# Controls false positive rate
sum(pValues < 0.05)
```

---

## Case study I: no true positives

```{r, dependson="createPvals"}
# Controls FWER 
sum(p.adjust(pValues,method="bonferroni") < 0.05)
# Controls FDR 
sum(p.adjust(pValues,method="BH") < 0.05)
```


---

## Case study II: 50% true positives

```{r createPvals2,cache=TRUE}
set.seed(1010093)
pValues <- rep(NA,1000)
for(i in 1:1000){
  x <- rnorm(20)
  # First 500 beta=0, last 500 beta=2
  if(i <= 500){y <- rnorm(20)}else{ y <- rnorm(20,mean=2*x)}
  pValues[i] <- summary(lm(y ~ x))$coeff[2,4]
}
trueStatus <- rep(c("zero","not zero"),each=500)
table(pValues < 0.05, trueStatus)
```

---


## Case study II: 50% true positives

```{r, dependson="createPvals2"}
# Controls FWER 
table(p.adjust(pValues,method="bonferroni") < 0.05,trueStatus)
# Controls FDR 
table(p.adjust(pValues,method="BH") < 0.05,trueStatus)
```


---


## Case study II: 50% true positives

__P-values versus adjusted P-values__
```{r, dependson="createPvals2",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(pValues,p.adjust(pValues,method="bonferroni"),pch=19)
plot(pValues,p.adjust(pValues,method="BH"),pch=19)
```


---


## Notes and resources

__Notes__:
* Multiple testing is an entire subfield
* A basic Bonferroni/BH correction is usually enough
* If there is strong dependence between tests there may be problems
  * Consider method="BY"

__Further resources__:
* [Multiple testing procedures with applications to genomics](http://www.amazon.com/Multiple-Procedures-Applications-Genomics-Statistics/dp/0387493166/ref=sr_1_2/102-3292576-129059?ie=UTF8&s=books&qid=1187394873&sr=1-2)
* [Statistical significance for genome-wide studies](http://www.pnas.org/content/100/16/9440.full)
* [Introduction to multiple testing](http://ies.ed.gov/ncee/pubs/20084018/app_b.asp)

