---
title       : Predicting with regression 
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Key ideas

* Fit a simple regression model
* Plug in new covariates and multiply by the coefficients
* Useful when the linear model is (nearly) correct

__Pros__:
* Easy to implement
* Easy to interpret

__Cons__:
* Often poor performance in nonlinear settings


---

## Example: Old faithful eruptions

<img class=center src=../../assets/img/08_PredictionAndMachineLearning/yellowstone.png height=400>

Image Credit/Copyright Wally Pacholka [http://www.astropics.com/](http://www.astropics.com/)

---

## Example: Old faithful eruptions

```{r faith}
library(caret);data(faithful); set.seed(333)
inTrain <- createDataPartition(y=faithful$waiting,
                              p=0.5, list=FALSE)
trainFaith <- faithful[inTrain,]; testFaith <- faithful[-inTrain,]
head(trainFaith)
```

---

## Eruption duration versus waiting time

```{r dependson="faith",fig.height=4,fig.width=4}
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
```

---

## Fit a linear model 

$$ ED_i = b_0 + b_1 WT_i + e_i $$

```{r faithlm,dependson="faith",fig.height=4,fig.width=4}
lm1 <- lm(eruptions ~ waiting,data=trainFaith)
summary(lm1)
```


---
## Model fit

```{r dependson="faithlm",fig.height=4,fig.width=4}
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(trainFaith$waiting,lm1$fitted,lwd=3)
```

---

## Predict a new value

$$\hat{ED} = \hat{b}_0 + \hat{b}_1 WT$$

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
coef(lm1)[1] + coef(lm1)[2]*80
newdata <- data.frame(waiting=80)
predict(lm1,newdata)
```

---

## Plot predictions - training and test

```{r ,dependson="faithlm",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(trainFaith$waiting,predict(lm1),lwd=3)
plot(testFaith$waiting,testFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration")
lines(testFaith$waiting,predict(lm1,newdata=testFaith),lwd=3)
```

---

## Get training set/test set errors

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
# Calculate RMSE on training
sqrt(sum((lm1$fitted-trainFaith$eruptions)^2))

# Calculate RMSE on test
sqrt(sum((predict(lm1,newdata=testFaith)-testFaith$eruptions)^2))
```

---

## Prediction intervals

```{r ,dependson="faithlm",fig.height=4,fig.width=4}
pred1 <- predict(lm1,newdata=testFaith,interval="prediction")
ord <- order(testFaith$waiting)
plot(testFaith$waiting,testFaith$eruptions,pch=19,col="blue")
matlines(testFaith$waiting[ord],pred1[ord,],type="l",,col=c(1,2,2),lty = c(1,1,1), lwd=3)
```


---

## Same process with caret

```{r caretfaith,dependson="faith",fig.height=4,fig.width=4}
modFit <- train(eruptions ~ waiting,data=trainFaith,method="lm")
summary(modFit$finalModel)
```


---

## Notes and further reading

* Regression models with multiple covariates can be included
* Often useful in combination with other models 
* [Elements of statistical learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
* [Modern applied statistics with S](http://www.amazon.com/Modern-Applied-Statistics-W-N-Venables/dp/0387954570)
* [Introduction to statistical learning](http://www-bcf.usc.edu/~gareth/ISL/)




