---
title       : Homework 2 for Stat Inference
subtitle    : (Use the arrow keys to navigate)
author      : Brian Caffo
job         : Johns Hopkins Bloomberg School of Public Health
framework   : io2012
highlighter : highlight.js  
hitheme     : tomorrow       
url:
    lib: ../../librariesNew #Remove new if using old slidify
    assets: ../../assets
widgets     : [mathjax, quiz, bootstrap]
mode        : selfcontained # {standalone, draft}
---
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, results='hide'}
# make this an external chunk that can be included in any file
library(knitr)
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## About these slides
- These are some practice problems for Statistical Inference Quiz 2
- They were created using slidify interactive which you will learn in 
Creating Data Products
- Please help improve this with pull requests here
(https://github.com/bcaffo/courses)

--- &radio
The probability that a manuscript gets accepted to a journal is 12% (say). However,
given that a revision is asked for, the probability that it gets accepted
is 90%. Is it possible that the probability that a manuscript has a revision
asked for is 20%? 

1. Yeah, that's totally possible.
2. _No, it's not possible._
3. It's not possible to answer this question.

*** .hint
$A = accepted$, $B = revision$. $P(A) = .12$, $P(A | B) = .90$. $P(B) = .20$

*** .explanation
$P(A \cap B) = P(A | B) * P(B) = .9 \times .2 = .18$ this is larger than
$P(A) = .12$, which is not possible since $A \cap B \subset A$.


--- &radio
Suppose that the number of web hits to a particular site are approximately normally
distributed with a mean of 100 hits per day and a standard deviation of 10 hits per day. What's the probability that a given day has fewer than 93 hits per day
expressed as a percentage to the nearest percentage point?

1. 76%
2. _24%_
3. 47%
4. 94%

*** .hint
Let $X$ be the number of hits per day. We want $P(X \leq 93)$ given that
$X$ is $N(100, 10^2)$.

*** .explanation
```{r}
round(pnorm(93, mean = 100, sd = 10) * 100)
```


--- &radio
Suppose 5% of housing projects have issues with asbestos. The sensitivity of a test
for asbestos is 93% and the specificity is 88%. What is the probability that a 
housing project has no asbestos given a negative test expressed as a percentage
to the nearest percentage point?

1. 0%
2. 5%
3. 10%
4. 20%
5. 50%
6. _100%_

*** .hint
$A = asbestos$, $T_+ = tests positive$, $T_- = tests negative$. 
$P(T_+ | A) = .93$, $P(T_- | A^c) = .88$, $P(A) = .05$.

*** .explanation
We want
$$
P(A^c | T_-) = \frac{P(T_- | A^c) P(A^c)}{P(T_- | A^c) P(A^c) + P(T_- | A) P(A)}
$$
```{r}
(.88 * .95) / (.88 * .95 + .07 * .05)
```



---  &multitext
Suppose that the number of web hits to a particular site are approximately normally
distributed with a mean of 100 hits per day and a standard deviation of 10 hits per day. 

1. What number of web hits per day represents the number so that only
5% of days have more hits? Express your answer to 3 decimal places.



*** .hint
Let $x$ be the number of hits per day. We want $x$ so that $F(x) = 0.95$.

*** .explanation
<span class="answer">`r round(qnorm(.95, mean = 100, sd = 10), 3)`</span>
```{r}
round(qnorm(.95, mean = 100, sd = 10), 3)
round(qnorm(.05, mean = 100, sd = 10, lower.tail = FALSE), 3)
```


---  &multitext
Suppose that the number of web hits to a particular site are approximately normally
distributed with a mean of 100 hits per day and a standard deviation of 10 hits per day. 

2. Imagine taking a random sample of 50 days. What number of web hits would
be the point so that only 5% of averages of 50 days of web traffic have more hits? 
Express your answer to 3 decimal places. 

*** .hint
Let $\bar X$ be the average number of hits per day for 50 randomly sampled days.
$X$ is $N(100, 10^2 / 50)$.

*** .explanation
<span class="answer">`r round(qnorm(.95, mean = 100, sd = 10 / sqrt(50) ), 3)`</span>
 
```{r}
round(qnorm(.95, mean = 100, sd = 10 / sqrt(50) ), 3)
round(qnorm(.05, mean = 100, sd = 10 / sqrt(50), lower.tail = FALSE), 3)
```

--- &multitext

You don't believe that your friend can discern good wine from cheap. Assuming
that you're right, in a blind test where you randomize 6 paired varieties (Merlot,
Chianti, ...) of cheap and expensive wines

what is the change that she gets 5 or 6 right expressed as a percentage
to one decimal place?

*** .hint
Let $p=.5$ and $X$ be binomial

*** .explanation

<span class="answer">`r round(pbinom(4, prob = .5, size = 6, lower.tail = FALSE) * 100, 1)`</span>

```{r}
round(pbinom(4, prob = .5, size = 6, lower.tail = FALSE) * 100, 1)
```

--- &multitext

Consider a uniform distribution. If we were to sample 100 draws from a 
a uniform distribution (which has mean 0.5, and variance 1/12) and take their
mean, $\bar X$

what is the approximate probability of getting as large as 0.51 or larger expressed to 3 decimal places?

*** .hint
Use the central limit theorem that says $\bar X \sim N(\mu, \sigma^2/n)$

*** .explanation

<span class="answer"> `r round(pnorm(.51, mean = 0.5, sd = sqrt(1 / 12 / 100), lower.tail = FALSE), 3)`</span>

```{r}
round(pnorm(.51, mean = 0.5, sd = sqrt(1 / 12 / 100), lower.tail = FALSE), 3)
```


--- &multitext

If you roll ten standard dice, take their average, then repeat this process over and over and construct a histogram, 

1. what would it be centered at?


*** .hint
$E[X_i] = E[\bar X]$ where $\bar X = \frac{1}{n}\sum_{i=1}^n X_i$

*** .explanation


The answer will be  <span class="answer">3.5</span> since the mean of the
sampling distribution of iid draws will be the population mean that the
individual draws were taken from.

--- &multitext

If you roll ten standard dice, take their average, then repeat this process over and over and construct a histogram, 

2. what would be its variance expressed to 3 decimal places?

*** .hint
$$Var(\bar X) = \sigma^2 /n$$

*** .explanation
The answer will be <span class="answer">`r round( mean( (1 : 6 - 3.5) ^2) / 10, 3)`</span> 
since the variance of the sampling distribution of the mean is $\sigma^2/10$
where $\sigma^2$ is the variance of a single die roll, which is 

```{r}
mean((1 : 6 - 3.5)^2 / 10)
```

--- &multitext
The number of web hits to a site is Poisson with mean 16.5 per day. 
What is the probability of getting 20 or fewer in 2 days expressed
as a percentage to one decimal place?

*** .hint
Let $X$ be the number of hits in 2 days then $X \sim Poisson(2\lambda)$

*** .explanation
<span class="answer">`r round(ppois(20, lambda = 16.5 * 2) * 100, 1)`</span>

```{r}
round(ppois(20, lambda = 16.5 * 2) * 100, 1)
```



