---
title       : Training options
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```


## SPAM Example

```{r loadPackage,cache=TRUE}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
modelFit <- train(type ~.,data=training, method="glm")
```

---

## Train options

```{r ,dependson="loadPackage"}
args(train.default)
```


---

## Metric options

__Continous outcomes__:
  * _RMSE_ = Root mean squared error
  * _RSquared_ = $R^2$ from regression models

__Categorical outcomes__:
  * _Accuracy_ = Fraction correct
  * _Kappa_ = A measure of [concordance](http://en.wikipedia.org/wiki/Cohen%27s_kappa)
  
  

--- 

## trainControl

```{r , dependson="loadPackage",cache=TRUE}
args(trainControl)
```

--- 

## trainControl resampling

* _method_
  * _boot_ = bootstrapping
  * _boot632_ = bootstrapping with adjustment
  * _cv_ = cross validation
  * _repeatedcv_ = repeated cross validation
  * _LOOCV_ = leave one out cross validation
* _number_
  * For boot/cross validation
  * Number of subsamples to take
* _repeats_
  * Number of times to repeate subsampling
  * If big this can _slow things down_


---

## Setting the seed

* It is often useful to set an overall seed
* You can also set a seed for each resample
* Seeding each resample is useful for parallel fits
* _seed_ must be a list with
  * Length equal to number of resamples
  * Length of each element equal to number of models fit



--- 



## seed example

```{r seedExample, dependson="loadPackage",cache=TRUE}
set.seed(1235); seeds <- vector(26,mode="list")
for(i in 1:26){seeds[[i]] <- floor(runif(1,0,1e5))}
trControl <- trainControl(seeds=seeds)
```



--- 


## seed example

```{r , dependson="seedExample",cache=TRUE}
modelFit2 <- train(type ~.,data=training, method="glm")
modelFit2
```


--- 

## seed example

```{r , dependson="seedExample",cache=TRUE}
modelFit3 <- train(type ~.,data=training, method="glm")
modelFit3
```

